---
title: "Benchmark results"
subtitle: "Which models do best on Multilingual GSM-Symbolic"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache.path = "_cache/benchmark_table/")
suppressPackageStartupMessages({ library(dplyr); library(tidyr); library(nanoparquet) })
```

A plain leaderboard, for readers who want one. Both splits are reported. English
is shown separately from the mean over the other 14 languages, and the gap
between them is the quantity the rest of the paper is about.

Everything here is built from the parquet, not from `model_cells.rds`, for two
reasons: `model_cells.rds` holds only the symbolic split, and it has already had
the analysis exclusions applied. The two runs excluded from the GLMM are
**reported here and marked** `*` -- a reader looking up a model should find it
even when it did not enter the fit.

Sampling error differs by split, because the two have different cell sizes: a
symbolic cell is 100 templates $\times$ 20 instances = 2,000 examples and an
original cell is 100 templates $\times$ 1 = 100. Per `sample_size.Rmd` that is
$\pm 2.3$ accuracy points per language on symbolic and $\pm 3.4$ on original.
A model's 15-language average is tighter, but only to about $\pm 1.8$ --- the
same 100 templates are used in every language, so the per-language errors are
correlated ($r = 0.67$), and independence would have given $\pm 0.6$.

```{r config}
EXCLUDED_LANGUAGES <- c("eng_metric", "uncorrected_isl", "nob")

# Runs kept in these tables but flagged, because they were dropped from the GLMM:
# both sit at the probability floor (0.28% and 0.40%) and exert large leverage on
# the logit scale. See the EXCLUDED_RUNS note in transfer_analysis.Rmd.
FLAGGED_RUNS <- c("Qwen3.5-0.8B (reasoning on)", "EuroLLM-1.7B-Instruct")
```

```{r data, cache=TRUE, cache.extra=list(EXCLUDED_LANGUAGES, FLAGGED_RUNS)}
raw <- nanoparquet::read_parquet("../../artifacts/transfer_tables/analysis.parquet") %>%
  filter(grepl("@refs/pr/16/", task, fixed = TRUE))

## The parquet's naming is NOT uniform, so the bare name cannot be assumed to
## mean one thing. Qwen3/Qwen3.5 label only the off run, making the bare name the
## reasoning-ON run; granite labels both explicitly; Olmo-3-Think appears only as
## "(reasoning on)". This reproduces transfer_analysis.Rmd's canonicalisation
## rather than re-deriving it -- getting it wrong silently mislabels the
## reasoning mode of every Qwen run.
off_bases <- raw %>% filter(grepl("\\(reasoning off\\)$", model)) %>%
  mutate(b = sub("\\s*\\(reasoning off\\)$", "", model)) %>% distinct(b) %>% pull(b)

cells <- raw %>%
  mutate(model = ifelse(model %in% off_bases, paste0(model, " (reasoning on)"), model),
         base_model = sub("\\s*\\(reasoning (on|off)\\)$", "", model),
         reasoning = ifelse(grepl("\\(reasoning on\\)$", model), "on", "off"),
         flagged = model %in% FLAGGED_RUNS) %>%
  filter(!language %in% EXCLUDED_LANGUAGES)

bylang <- cells %>%
  group_by(model, base_model, reasoning, flagged, params_b, family, language, split) %>%
  summarise(acc = 100 * mean(correct), n = n(), .groups = "drop")

cat(n_distinct(bylang$model), "runs x", n_distinct(bylang$language), "languages x",
    n_distinct(bylang$split), "splits =", nrow(bylang), "rows\n")
cat("flagged runs present:", sum(unique(bylang[c("model","flagged")])$flagged), "\n")
cat("norwegian present?", "nob" %in% bylang$language, "\n")
cat("examples per cell:", paste(range(bylang$n), collapse = "-"), "\n")
```

# Overview

Ordered by symbolic accuracy averaged over the 15 languages.

```{r overview, cache=TRUE, dependson='data'}
tab <- bylang %>% filter(split == "synthetic") %>%
  group_by(model, base_model, reasoning, flagged, params_b, family) %>%
  summarise(overall = mean(acc),
            english = acc[language == "eng"],
            non_english = mean(acc[language != "eng"]),
            worst = min(acc),
            worst_lang = as.character(language[which.min(acc)]),
            .groups = "drop") %>%
  mutate(gap = english - non_english) %>%
  arrange(desc(overall))

## Original-split overall, for the leaderboard's second accuracy column.
orig <- bylang %>% filter(split == "original") %>%
  group_by(model) %>% summarise(overall_orig = mean(acc), .groups = "drop")
tab <- tab %>% left_join(orig, by = "model")

tab %>% transmute(Model = base_model, Reas. = reasoning, `Params (B)` = params_b,
                  Original = round(overall_orig, 1), Symbolic = round(overall, 1),
                  English = round(english, 1), `Non-English` = round(non_english, 1),
                  Gap = round(gap, 1),
                  Worst = paste0(round(worst, 1), " (", worst_lang, ")"),
                  Fit = ifelse(flagged, "excluded", ""))
```

```{r overview-latex, results='asis'}
esc <- function(x) gsub("_", "\\\\_", x)
dir.create("tables", showWarnings = FALSE)
sink("tables/leaderboard.tex", split = TRUE)   # split: also echo into the knit
cat("\\begin{table}[p]\n\\centering\n\\small\n")
cat("\\begin{tabular}{llrrrrrrl}\n\\toprule\n")
cat("Model & Reas. & Params & Orig. & Symb. & English & Non-Eng. & Gap & Worst \\\\\n")
cat("\\midrule\n")
for (i in seq_len(nrow(tab))) with(tab[i, ], cat(sprintf(
  "%s%s & %s & %.1f & %.1f & %.1f & %.1f & %.1f & %.1f & %.1f (%s) \\\\\n",
  esc(base_model), ifelse(flagged, "$^{*}$", ""), reasoning, params_b,
  overall_orig, overall, english, non_english, gap, worst, worst_lang)))
cat("\\bottomrule\n\\end{tabular}\n")
cat("\\caption{Exact-answer accuracy on Multilingual GSM-Symbolic, averaged over the 15",
    "benchmark languages and ordered by symbolic accuracy. \\emph{Gap} is English minus the",
    "mean of the other 14 languages. \\emph{Worst} is the lowest-scoring language on the",
    "symbolic split. Per-language scores carry roughly $\\pm 2.3$ points of sampling error",
    "on the symbolic split and $\\pm 3.4$ on the original; the 15-language averages about",
    "$\\pm 1.8$ (Appendix~\\ref{app:sec:sample-size}).",
    "$^{*}$Reported here but excluded from the model in Equation~\\ref{eq:model}.}\n")
cat("\\label{tab:leaderboard}\n\\end{table}\n")
sink()
```

# Summary

```{r summary}
cat(sprintf("best overall: %s (%.1f%%)\n", tab$model[1], tab$overall[1]))
cat(sprintf("largest English-to-non-English gap: %s (%.1f points)\n",
            tab$model[which.max(tab$gap)], max(tab$gap)))

## Apertus is the one model whose English score is far BELOW its non-English
## scores, so its "gap" is large and negative. That is a known anomaly rather
## than good transfer, and reporting the minimum gap without saying so would
## crown it the most balanced model in the suite.
neg <- tab %>% filter(gap < 0)
if (nrow(neg)) {
  cat("\nruns scoring WORSE in English than in the average other language:\n")
  print(neg %>% transmute(model, english = round(english, 1),
                          non_english = round(non_english, 1), gap = round(gap, 1)))
}

paired <- tab %>% filter(!flagged) %>%
  select(base_model, reasoning, overall, gap) %>%
  pivot_wider(names_from = reasoning, values_from = c(overall, gap)) %>%
  filter(!is.na(overall_on), !is.na(overall_off))
cat(sprintf("\nacross %d base models run both ways, reasoning adds %.1f points overall\n",
            nrow(paired), mean(paired$overall_on - paired$overall_off)))
cat(sprintf("and narrows the English gap by %.1f points\n",
            mean(paired$gap_off - paired$gap_on)))

cat(sprintf("\nsymbolic is below original by %.1f points on average\n",
            mean(tab$overall_orig - tab$overall)))
```

# Results by language

A 720-row long table is unreadable on the page, so this goes out as two
run $\times$ language matrices, one per split. Within a split the per-cell
sampling error is near-uniform, so intervals add nothing that one caption
sentence cannot carry -- but the two splits differ ($\pm 2.3$ vs $\pm 3.4$),
so each caption states its own.

```{r by-language-latex, results='asis'}
langs <- sort(unique(as.character(bylang$language)))
ord <- tab %>% transmute(model, overall)
stopifnot(setequal(ord$model, unique(bylang$model)))

emit <- function(sp, label, cap, file) {
  sink(file.path("tables", file), split = TRUE)
  m <- bylang %>% filter(split == sp) %>%
    select(model, base_model, reasoning, flagged, language, acc) %>%
    tidyr::pivot_wider(names_from = language, values_from = acc) %>%
    left_join(ord, by = "model") %>%
    arrange(desc(overall))
  cat("\\begin{sidewaystable}\n\\centering\n\\resizebox{\\textheight}{!}{%\n")
  cat("\\begin{tabular}{ll", strrep("r", length(langs)), "}\n\\toprule\n", sep = "")
  cat("Model & Reas. & ", paste(langs, collapse = " & "), " \\\\\n\\midrule\n", sep = "")
  for (i in seq_len(nrow(m))) {
    cat(esc(m$base_model[i]), if (m$flagged[i]) "$^{*}$" else "", " & ",
        m$reasoning[i], " & ",
        paste(sprintf("%.1f", unlist(m[i, langs])), collapse = " & "), " \\\\\n", sep = "")
  }
  cat("\\bottomrule\n\\end{tabular}}\n")
  cat("\\caption{", cap, "}\n\\label{", label, "}\n\\end{sidewaystable}\n\n", sep = "")
  sink()
}

emit("original", "tab:by-language-original",
     paste("Exact-answer accuracy (\\%) on the \\textbf{original} split, by run and language.",
           "Rows are ordered by mean symbolic accuracy, matching",
           "Table~\\ref{tab:leaderboard}. Each cell is 100 questions, one per template, and",
           "carries roughly $\\pm 3.4$ points of sampling error",
           "(Appendix~\\ref{app:sec:sample-size}).",
           "$^{*}$Excluded from the model in Equation~\\ref{eq:model}."),
     "by_language_original.tex")
emit("synthetic", "tab:by-language-symbolic",
     paste("Exact-answer accuracy (\\%) on the \\textbf{symbolic} split, by run and language.",
           "Row order and $^{*}$ as in Table~\\ref{tab:by-language-original}. Each cell is",
           "2{,}000 questions, 20 per template, and carries roughly $\\pm 2.3$ points of",
           "sampling error."),
     "by_language_symbolic.tex")
```
