---
title: "Answer-format compliance"
subtitle: "Does the model emit \\boxed{} less often in some languages?"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache.path = "_cache/format_compliance/")
suppressPackageStartupMessages({ library(dplyr); library(lme4) })
```

The prompt asks for the final answer in `\boxed{}`. If models comply less often
in some languages than others, part of the transfer gap could be a formatting
artefact rather than a reasoning failure. This document checks that.

Two measurements, and the difference between them is the point:

- **`has_boxed`** -- did the completion contain `\boxed{`, the format asked for.
- **`extracted`** -- could the scorer recover *any* answer. It also accepts the
  GSM8K `#### n` form, so a missing `\boxed` is not automatically a lost point.
  This is the one that can cost accuracy.

Data comes from `collect_compliance.py`, which parses all 958 synthetic eval
logs on the Hub (`refs/pr/16`) and writes `compliance.csv`. It reads every 4th
sample, 500 of 2,000 per log, which is ample for a rate.

```{r data, cache=TRUE}
d <- read.csv("../../artifacts/analysis/compliance.csv", stringsAsFactors = FALSE) %>%
  mutate(model = sub("^.*/", "", model)) %>%     # vllm/org/name -> name
  group_by(model, language) %>%
  summarise(across(c(n, has_boxed, extracted, correct), sum), logs = n(),
            .groups = "drop") %>%
  mutate(boxed_rate = 100 * has_boxed / n,
         ext_rate   = 100 * extracted / n,
         acc        = 100 * correct / n)

cat("cells:", nrow(d), "=", n_distinct(d$model), "models x",
    n_distinct(d$language), "languages\n")
cat("questions parsed:", format(sum(d$n), big.mark = ","), "\n")
stopifnot(nrow(d) == n_distinct(d$model) * n_distinct(d$language))  # complete grid
```

A cell pools the logs belonging to it. Most cells hold one or two; the extras
are reruns and reasoning variants. **The eval log's model field does not record
the reasoning mode**, so a base model's reasoning-on and reasoning-off runs are
pooled here and this document works with 36 base models rather than the paper's
48 runs. That is a limitation for reading any single row, not for the language
question, which is asked within model.

# Per-language rates

Averaged over models rather than pooled over logs: pooling would weight a model
by how many times it happened to be rerun.

```{r per-language}
lang <- d %>% group_by(language) %>%
  summarise(boxed = mean(boxed_rate), ext = mean(ext_rate), acc = mean(acc),
            .groups = "drop") %>%
  mutate(recovered = ext - boxed) %>%
  arrange(desc(boxed))
lang %>% mutate(across(where(is.numeric), ~round(.x, 1))) %>% as.data.frame()
```

Compliance does vary, from `r sprintf("%.1f%%", max(lang$boxed))` to
`r sprintf("%.1f%%", min(lang$boxed))`, and it is not an English-versus-rest
split: German and Italian comply *more* often than English.

# Is the language variation real?

The grid is complete, so this is a within-model comparison -- each model
contributes a deviation from its own compliance rate.

```{r glmm, cache=TRUE, dependson='data'}
ctrl <- glmerControl(optimizer = "bobyqa")
f1 <- glmer(cbind(has_boxed, n - has_boxed) ~ (1|model) + (1|language) + (1|model:language),
            d, binomial, control = ctrl)
f0 <- glmer(cbind(has_boxed, n - has_boxed) ~ (1|model) + (1|model:language),
            d, binomial, control = ctrl)
print(as.data.frame(VarCorr(f1)) %>% transmute(group = grp, sd = round(sdcor, 3)) %>%
        arrange(desc(sd)))
anova(f0, f1)
```

Yes, decisively -- but it is much the smaller term. Which model you use matters
about four times as much as which language you ask in.

# Is it a cause of the gap, or a symptom?

Compliance correlates with accuracy across languages at
`r sprintf("%.2f", cor(lang$boxed, lang$acc))`, which invites the reading that
poor formatting costs accuracy. The causal arrow is more likely the reverse: a
model that cannot solve the problem rambles and never reaches a boxed answer.

The two readings make opposite predictions for models that *can* solve the
problem. If formatting were a language-specific handicap it should persist in
strong models; if it is a symptom of failure it should vanish.

```{r strong-weak, cache=TRUE, dependson='data'}
ms <- d %>% group_by(model) %>% summarise(m = mean(acc), .groups = "drop")
strong <- ms %>% filter(m >= 85) %>% pull(model)
weak   <- ms %>% filter(m <  40) %>% pull(model)

summarise_set <- function(models, label) {
  d %>% filter(model %in% models) %>% group_by(language) %>%
    summarise(boxed = round(mean(boxed_rate), 1), ext = round(mean(ext_rate), 1),
              acc = round(mean(acc), 1), .groups = "drop") %>%
    arrange(desc(boxed)) %>% mutate(set = label)
}
cat(length(strong), "strong models (mean acc >= 85%),", length(weak), "weak (< 40%)\n")
as.data.frame(summarise_set(strong, "strong"))
as.data.frame(summarise_set(weak, "weak"))
```

```{r strong-detail}
s <- d %>% filter(model %in% strong)
cat(sprintf("strong: mean compliance %.2f%%, worst single cell %.2f%% (%s / %s)\n",
            mean(s$boxed_rate), min(s$boxed_rate),
            s$model[which.min(s$boxed_rate)], s$language[which.min(s$boxed_rate)]))
cat(sprintf("strong: spread across the 15 language means = %.2f pp\n",
            diff(range(tapply(s$boxed_rate, s$language, mean)))))
cat(sprintf("strong: English %.1f%% vs Marathi %.1f%% accuracy, a %.1f pp gap\n",
            mean(s$acc[s$language == "eng"]), mean(s$acc[s$language == "mar"]),
            mean(s$acc[s$language == "eng"]) - mean(s$acc[s$language == "mar"])))
```

Symptom. Among the strong models compliance is essentially perfect in every
language -- the whole spread is a fraction of a point -- while the accuracy gap
between English and Marathi is still 16.8 points. The collapse in the weak
models is real but it is a collapse in models that are failing anyway.

# Ceiling on the format explanation

The generous version of the question: if every answer the scorer could not
extract would otherwise have been correct, how much of each language's gap to
English would that explain?

```{r ceiling}
eng <- lang %>% filter(language == "eng")
lang %>% filter(language != "eng") %>%
  transmute(language,
            acc_gap = round(eng$acc - acc, 1),
            ext_gap = round(eng$ext - ext, 1),
            max_share_pct = round(100 * (eng$ext - ext) / (eng$acc - acc))) %>%
  arrange(desc(acc_gap)) %>% as.data.frame()
```

At most 11--16% in the four worst languages, and negative in the eight languages
whose answers extract *more* reliably than English's do. The extraction fallback is why: it recovers
most of the missing `\boxed`, and recovers most where compliance is worst
(Marathi +10.5 points, Icelandic +7.3, against +0.2 in English).
