---
title: "Model ablations: the template-by-language random effect"
subtitle: "Appendix A.1"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache.path = "_cache/model_ablations/")
suppressPackageStartupMessages({ library(dplyr); library(lme4) })
```

The main model carries `(1 | template)` but no template-by-language term, i.e. it
assumes a template's difficulty is the same in every language up to that
language's overall effect. It is not: decomposing per-(language, template)
accuracy on the logit scale gives sd(template) around 0.69 against
sd(template x language) around 0.52, and the cross-language Spearman correlation
of template difficulty averages 0.73.

Two things to establish:

1. what the omitted component does to the fixed effects and their standard
   errors --- in particular the language-level terms, which are the paper's
   headline estimates;
2. whether the variance decomposition reported in the results is stable, since
   those percentages are a *ratio between two fits* and adding the term changes
   both.

The other two ablations in Appendix A.1 --- the resource x distance interaction
and script as a feature --- live in `transfer_analysis.Rmd`, which already fits
them alongside the main model.

```{r data, cache=TRUE}
cells <- readRDS("../../artifacts/analysis/model_cells.rds")
ctrl <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 5e5))

RE_OLD <- paste("(1 | family/base_model) + (1 | model) + (1 | template) +",
                "(1 | language) + (1 | model:language)")
RE_NEW <- paste(RE_OLD, "+ (1 | template:language)")
FEAT <- "(relative_fertility_within_z + typological_distance_z + log_resource_z)"
rhs <- function(RE) paste(FEAT, "* log_params_z +",
                          "reasoning * (relative_fertility_within_z +",
                          "typological_distance_z + log_resource_z + log_params_z) +", RE)
Y <- "cbind(n_correct, n_total - n_correct) ~"
cat("cells:", nrow(cells), "| languages:", nlevels(cells$language),
    "| template:language levels:",
    nlevels(interaction(cells$template, cells$language, drop = TRUE)), "\n")
```

# Fits

Both the full model and the size-only baseline are refitted with the extra term,
because the variance decomposition compares the two.

```{r fit, cache=TRUE, dependson='data', cache.extra=RE_NEW}
fit_tl <- glmer(as.formula(paste(Y, rhs(RE_NEW))), data = cells,
                family = binomial(), control = ctrl)
base_tl <- glmer(as.formula(paste(Y, "log_params_z + reasoning +", RE_NEW)),
                 data = cells, family = binomial(), control = ctrl)
```

The models the paper reports come out of `transfer_analysis.Rmd`'s cache, so this
document always compares against exactly what that document printed.

```{r reference-fits, cache=TRUE, dependson='data'}
cache <- function(pat) {
  f <- list.files("_cache/transfer_analysis", pattern = pat, full.names = TRUE)
  f <- f[grepl("\\.RData$", f)]
  stopifnot(length(f) >= 1)
  e <- new.env(); load(f[order(file.mtime(f), decreasing = TRUE)][1], envir = e); e
}
fit_old  <- cache("^fit-lme4_")$fit_main
base_old <- cache("^fit-baseline_")$fit_baseline
```

```{r convergence}
for (nm in c("fit_tl", "base_tl")) {
  f <- get(nm)
  cat(nm, "| conv:",
      if (length(f@optinfo$conv$lme4$messages)) f@optinfo$conv$lme4$messages else "OK", "\n")
}
```

# Variance components

```{r variance-components}
sdl <- function(f, g) { v <- VarCorr(f); if (is.null(v[[g]])) NA_real_ else as.numeric(attr(v[[g]], "stddev")[1]) }
gs <- unique(c(names(VarCorr(fit_old)), names(VarCorr(fit_tl))))
data.frame(component = gs,
           reported = round(vapply(gs, sdl, numeric(1), f = fit_old), 4),
           with_term = round(vapply(gs, sdl, numeric(1), f = fit_tl), 4)) %>%
  arrange(desc(coalesce(with_term, 0)))
```

# Fixed effects

The question is not whether the term belongs --- it plainly does --- but whether
leaving it out changed anything we report.

```{r fixed-effects}
co_o <- summary(fit_old)$coefficients; co_n <- summary(fit_tl)$coefficients
tt <- rownames(co_o)
data.frame(term = tt,
           est_reported = round(co_o[tt, 1], 4), est_with_term = round(co_n[tt, 1], 4),
           se_reported  = round(co_o[tt, 2], 4), se_with_term  = round(co_n[tt, 2], 4),
           se_ratio = round(co_n[tt, 2] / co_o[tt, 2], 2),
           p_with_term = signif(co_n[tt, 4], 3))
```

```{r sign-and-significance}
sig_o <- co_o[, 4] < .05; sig_n <- co_n[, 4] < .05
cat("significant in both:", sum(sig_o & sig_n), "of", sum(sig_o), "reported as significant\n")
cat("significant terms keeping their sign:",
    sum(sign(co_o[sig_o, 1]) == sign(co_n[sig_o, 1])), "of", sum(sig_o), "\n")
cat("terms changing sign at all:", sum(sign(co_o[, 1]) != sign(co_n[, 1])),
    "(check whether these are null in both)\n")
print(round(cbind(reported = co_o[sign(co_o[, 1]) != sign(co_n[, 1]), 1, drop = FALSE],
                  with_term = co_n[sign(co_o[, 1]) != sign(co_n[, 1]), 1, drop = FALSE]), 4))
cat("\nAIC reported / with term:", round(AIC(fit_old), 1), "/", round(AIC(fit_tl), 1), "\n")
```

# Is the variance decomposition stable?

The results section quotes the share of each component the language features
remove, relative to a size-only baseline. Both fits change here, so the ratio has
to be recomputed rather than inferred from the coefficients being stable.

```{r variance-explained}
sp <- function(b, f, g) 1 - sdl(f, g)^2 / sdl(b, g)^2
comp <- c("language", "model:language")
data.frame(component = comp,
           reported = round(100 * sapply(comp, function(g) sp(base_old, fit_old, g)), 1),
           with_term = round(100 * sapply(comp, function(g) sp(base_tl, fit_tl, g)), 1))
```

```{r variance-combined}
cmb <- function(b, f) 1 - sum(sapply(comp, function(g) sdl(f, g)^2)) /
                          sum(sapply(comp, function(g) sdl(b, g)^2))
cat(sprintf("combined: reported %.1f%%, with term %.1f%%\n",
            100 * cmb(base_old, fit_old), 100 * cmb(base_tl, fit_tl)))
cat(sprintf("sd(template:language) in the baseline %.4f vs the full model %.4f\n",
            sdl(base_tl, "template:language"), sdl(fit_tl, "template:language")))
cat("-- near-identical means the language features explain none of it, which is\n")
cat("   what a genuine nuisance component should look like.\n")
```
