---
title: "Predicting performance on an unseen language"
subtitle: "Section 4.4 and Appendix E"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache.path = "_cache/predict_new_language/")
suppressPackageStartupMessages({ library(dplyr); library(tidyr); library(lme4); library(parallel) })
```

Two questions, both answered by holding each language out in turn:

1. **With no evaluation at all**, how well can we forecast a language we have
   never run? Every predictor is obtainable without touching the target
   language, so this is a forecast rather than a fit statistic.
2. **With a little evaluation**, how much does running $k$ templates in the
   target language buy?

```{r shared, cache=TRUE}
cells <- readRDS("../../artifacts/analysis/model_cells.rds")
out_dir <- "../../artifacts/analysis"
ctrl <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 3e5))

RE <- paste("(1 | family/base_model) + (1 | model) + (1 | template) +",
            "(1 | language) + (1 | model:language)")
FEAT <- "(relative_fertility_within_z + typological_distance_z + log_resource_z)"
RHS <- paste(FEAT, "* log_params_z +",
             "reasoning * (relative_fertility_within_z +",
             "typological_distance_z + log_resource_z + log_params_z) +", RE)
Y <- "cbind(n_correct, n_total - n_correct) ~"

## Standardise on the training fold only, then apply those constants to the
## held-out language, so the target's own mean and spread never leak in.
apply_z <- function(train, test) {
  for (p in list(c("relative_fertility_within", "relative_fertility_within_z"),
                 c("typological_distance_from_english", "typological_distance_z"),
                 c("log10_common_crawl_pages", "log_resource_z"),
                 c("log2_params", "log_params_z"))) {
    s <- c(mean(train[[p[1]]]), sd(train[[p[1]]]))
    train[[p[2]]] <- (train[[p[1]]] - s[1]) / s[2]
    test[[p[2]]]  <- (test[[p[1]]] - s[1]) / s[2]
  }
  list(train = train, test = test)
}

langs <- levels(droplevels(cells$language))
truth <- cells %>% group_by(language, model) %>%
  summarise(truth = sum(n_correct) / sum(n_total), .groups = "drop")
cat("languages:", length(langs), "|", paste(langs, collapse = " "), "\n")
```

# No evaluation at all

Prediction excludes the `language` and `model:language` random effects, which are
by definition unknown for a language never evaluated. The model, family and
template effects *are* known from the other languages and are kept. The excluded
components are therefore irreducible error --- which is exactly what the variance
decomposition in the results predicts.

Baselines, in increasing order of what a practitioner would actually do:

| baseline | is |
|---|---|
| `grand_mean` | the training-set accuracy, ignoring everything |
| `model_mean` | this model's average over the languages already evaluated --- the one the feature model has to beat to be worth anything |
| `size_only` | a fitted model with size but no language features |
| `features` | the full model |

```{r lolo-fits}
## Each fold runs as its own R process (run_fold.R). R does not return heap freed
## by glmer to the OS, so fitting 15 folds in one session grows memory until the
## machine dies -- rm() and gc() are not enough. One process per fold also makes
## the job resumable: a completed fold is a file in artifacts/analysis/folds/.
FOLDS <- "../../artifacts/analysis/folds"
LOLO_CSV <- file.path(out_dir, "lolo_predictions.csv")

if (!file.exists(LOLO_CSV)) {
  for (L in langs) {
    if (file.exists(file.path(FOLDS, paste0("lolo_", L, ".rds")))) next
    cat("fitting fold", L, "... "); flush.console()
    t0 <- Sys.time()
    system2("Rscript", c("run_fold.R", "lolo", L))
    cat(round(as.numeric(difftime(Sys.time(), t0, units = "mins")), 1), "min\n")
  }
  got <- list.files(FOLDS, pattern = "^lolo_.*[.]rds$", full.names = TRUE)
  stopifnot(length(got) == length(langs))
  write.csv(bind_rows(lapply(got, readRDS)), LOLO_CSV, row.names = FALSE)
}
pred <- read.csv(LOLO_CSV)
cat(nrow(pred), "rows,", n_distinct(pred$language), "languages\n")
```

```{r lolo-scores}
METHODS <- c("features", "size_only", "model_mean", "grand_mean")
score <- function(d, by) {
  d %>% group_by(across(all_of(by))) %>%
    summarise(across(all_of(c("observed", METHODS)), ~ weighted.mean(., n_total)), .groups = "drop") %>%
    pivot_longer(all_of(METHODS), names_to = "method", values_to = "pred") %>%
    group_by(method) %>%
    summarise(MAE = 100 * mean(abs(pred - observed)),
              RMSE = 100 * sqrt(mean((pred - observed)^2)),
              pearson = cor(pred, observed),
              spearman = cor(pred, observed, method = "spearman"), .groups = "drop") %>%
    arrange(MAE) %>% mutate(across(where(is.numeric), ~round(., 3)))
}
```

Predicting a **novel language's mean difficulty**. Note that `model_mean` and
`grand_mean` coincide here: averaging each model's own training mean over all
models recovers the overall training mean, so the per-model baseline does no work
at this level.

```{r lolo-language}
score(pred, "language")
```

Predicting a **specific model in a novel language**, where the per-model baseline
does separate from the grand mean:

```{r lolo-model}
score(pred, c("language", "model"))
```

```{r lolo-per-language}
pred %>% group_by(language) %>%
  summarise(observed = weighted.mean(observed, n_total),
            predicted = weighted.mean(features, n_total), .groups = "drop") %>%
  mutate(error_pp = 100 * (predicted - observed)) %>%
  arrange(error_pp) %>% mutate(across(where(is.numeric), ~round(., 3)))
```

# With a few evaluated templates

Now the $k$ chosen templates from the target language are **added to the training
data and the model is refitted from scratch**. Nothing is frozen, so there is no
approximation to defend --- only compute.

Design notes:

- **nested draws**: within a replicate the template sets are prefixes of one
  shuffle ($S_1 \subset S_2 \subset \dots \subset S_{20}$), so movement along the
  curve is not confounded with which templates were drawn;
- **replicates**: `R` independent shuffles per language, since a single draw of
  one template is extremely noisy;
- **scoring**: primary is the $100 - k$ templates *not* used in training, which
  is the honest forecast. The all-100 figure is reported too, since scoring on
  templates you trained on flatters large $k$;
- **$k = 0$** is the leave-one-language-out fit above.

The specification is the paper's, with no template-by-language term. That matters
at small $k$: with no such term the model must charge a single template's
idiosyncrasy to the language effect, so very small $k$ can be *worse* than not
evaluating at all. That is a property of the model rather than of the method (see
`model_ablations.Rmd`), and it is reported rather than smoothed over --- watch
`sd_language` inflate at $k = 1$.

```{r fewshot-fits}
KS <- c(1, 2, 5, 10, 20)
R <- 3                # replicates per (language, k). One draw is not enough at
                      # small k: a single unlucky template set moved the
                      # 15-language mean by over a point. Replicates also give a
                      # handle on how much the template draw matters at all,
                      # separately from which languages we happen to have.
WORKERS <- 2          # 16 GB; each glmer peaks well over 1 GB
FEW_CSV <- file.path(out_dir, "few_shot_refit.csv")

## Jobs are ordered so the most-cited k finishes first: if the run is interrupted
## the curve is still usable at k = 10, then 5, 20, 2, 1.
jobs <- expand.grid(rep = seq_len(R), language = langs, k = c(10, 5, 20, 2, 1),
                    stringsAsFactors = FALSE)[, c("language", "k", "rep")]
jobs <- jobs[order(match(jobs$k, c(10, 5, 20, 2, 1))), ]
cat(nrow(jobs), "fits (", length(langs), "languages x", R, "replicates x",
    length(KS), "k-values ),", WORKERS, "at a time\n")

if (!file.exists(FEW_CSV)) {
  todo <- jobs[!file.exists(file.path(FOLDS, sprintf("few_%s_k%02d_r%d.rds",
                                                     jobs$language, jobs$k, jobs$rep))), ]
  cat(nrow(todo), "still to run\n"); flush.console()
  if (nrow(todo)) {
    ## xargs gives one fresh process per fit, for the same reason as above
    spec <- paste(todo$language, todo$k, todo$rep)
    writeLines(spec, tf <- tempfile())
    system2("xargs", c("-P", WORKERS, "-L", "1",
                       "Rscript", "run_fold.R", "fewshot"), stdin = tf)
  }
  ## Check the jobs we asked for, not the file count: earlier, wider grids leave
  ## extra replicates on disk, and counting those would never match nrow(jobs).
  want <- file.path(FOLDS, sprintf("few_%s_k%02d_r%d.rds", jobs$language, jobs$k, jobs$rep))
  cat("completed", sum(file.exists(want)), "of", nrow(jobs), "requested\n")
  if (all(file.exists(want))) {
    ## keep any extra replicates that already exist -- they are free precision
    all_f <- list.files(FOLDS, pattern = "^few_.*[.]rds$", full.names = TRUE)
    write.csv(bind_rows(lapply(all_f, readRDS)), FEW_CSV, row.names = FALSE)
  }
}
few <- if (file.exists(FEW_CSV)) read.csv(FEW_CSV) else
  bind_rows(lapply(list.files(FOLDS, pattern = "^few_.*[.]rds$", full.names = TRUE), readRDS))
cat("fits available:", nrow(few), "\n")
```

```{r fewshot-curve}
b <- pred %>% group_by(language, model) %>%
  summarise(truth = sum(n_correct) / sum(n_total),
            feat = weighted.mean(features, n_total), .groups = "drop") %>%
  group_by(language) %>%
  summarise(mae = mean(abs(feat - truth)), err = mean(feat) - mean(truth), .groups = "drop")

## Two sources of uncertainty, and they answer different questions.
##
##   across languages  how much would the curve move with a different set of 15
##                     languages? This is the one that belongs on the reported
##                     numbers, and it applies at every k including k = 0.
##   across draws      how much does the particular template set matter at a
##                     given k? Only defined for k > 0, and it is a property of
##                     the method rather than of our estimate of it.
##
## The bootstrap resamples LANGUAGES (averaging replicates within a language
## first), so it is paired across k: the same languages are drawn for every
## budget, which is the right comparison since each k is evaluated on all 15.
set.seed(7)
BOOT <- 2000
per_lang <- few %>% group_by(k, language) %>%
  summarise(mae = mean(mae_heldout), err = mean(abs(err_heldout)),
            n_draws = n(), .groups = "drop") %>%
  bind_rows(b %>% transmute(k = 0, language, mae, err = abs(err), n_draws = 1L))

boot_ci <- function(x) {
  m <- replicate(BOOT, mean(sample(x, length(x), replace = TRUE)))
  c(lo = unname(quantile(m, .025)), hi = unname(quantile(m, .975)))
}
curve <- per_lang %>% group_by(k) %>%
  summarise(model_in_language = 100 * mean(mae),
            ml_lo = 100 * boot_ci(mae)[1], ml_hi = 100 * boot_ci(mae)[2],
            language_mean = 100 * mean(err),
            lm_lo = 100 * boot_ci(err)[1], lm_hi = 100 * boot_ci(err)[2],
            draws = first(n_draws), .groups = "drop") %>% arrange(k)

curve %>% transmute(k, draws,
  `model in language` = sprintf("%.2f [%.2f, %.2f]", model_in_language, ml_lo, ml_hi),
  `language mean` = sprintf("%.2f [%.2f, %.2f]", language_mean, lm_lo, lm_hi))
```

Brackets are 95% bootstrap intervals over the 15 languages. They are wide because
15 is a small sample of languages, and they are *paired* across $k$ --- the same
languages are resampled at every budget --- so the intervals overlapping does not
mean the differences between budgets are unreliable. The paired comparison below
is the right test for that.

```{r fewshot-paired}
## Does a budget beat k = 0 on the SAME languages? Paired differences remove the
## between-language variation that dominates the intervals above.
base0 <- per_lang %>% filter(k == 0) %>% select(language, mae0 = mae, err0 = err)
few %>% group_by(k, language) %>%
  summarise(mae = mean(mae_heldout), err = mean(abs(err_heldout)), .groups = "drop") %>%
  inner_join(base0, by = "language") %>%
  group_by(k) %>%
  summarise(n = n(),
            d_model = 100 * mean(mae - mae0),
            p_model = suppressWarnings(wilcox.test(mae, mae0, paired = TRUE)$p.value),
            d_lang = 100 * mean(err - err0),
            p_lang = suppressWarnings(wilcox.test(err, err0, paired = TRUE)$p.value),
            .groups = "drop") %>%
  mutate(across(where(is.numeric), ~round(., 3)))
```

```{r fewshot-draws}
## How much does the template draw matter, at each k? This is the spread the
## replicates exist to measure.
few %>% group_by(k, language) %>%
  summarise(sd_draw = sd(100 * mae_heldout), .groups = "drop") %>%
  group_by(k) %>%
  summarise(mean_sd_across_draws = round(mean(sd_draw, na.rm = TRUE), 2),
            worst_language = round(max(sd_draw, na.rm = TRUE), 2), .groups = "drop")
```

`sd(language)` by $k$. Inflation at small $k$ is the model charging a single
template's idiosyncrasy to the language effect:

```{r fewshot-sd}
few %>% group_by(k) %>%
  summarise(fits = n(), languages = n_distinct(language),
            sd_language = round(mean(sd_language), 4), .groups = "drop")
```
