---
title: "The symbolic penalty"
subtitle: "Appendix C.1 -- are symbolic variants harder, and does it vary by language?"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache.path = "_cache/symbolic_penalty/")
suppressPackageStartupMessages({ library(dplyr); library(lme4); library(nanoparquet) })
```

The descriptive claim --- synthetic variants are harder --- is a paired comparison
over 768 model x language gaps, which ignores the nested structure. Here the
penalty is fitted properly: both splits in one binomial GLMM, with `split` as a
fixed effect and as a random slope over language.

| term | asks |
|---|---|
| `split` | the average symbolic penalty |
| `split x` language features | is the penalty predicted by resource level or typological distance? |
| `(1 + split \| language)` | does the penalty vary by language beyond what the features explain? |

The likelihood-ratio test of the random slope against a random-intercept-only
model is the test of *does fragility differ across languages at all*.

# Data

The exclusions match `transfer_analysis.Rmd`: `eng_metric` and `uncorrected_isl`
are diagnostic labels rather than languages, and `nob` is not one of the
benchmark's 15.

Note the two splits differ in instances per cell --- the original split has one
instance per (model, language, template), the synthetic split has 20. The
binomial outcome handles this; it means the original arm simply carries less
information per cell.

```{r prep, cache=TRUE}
EXCLUDED_LANGUAGES <- c("eng_metric", "uncorrected_isl", "nob")
## Read straight from the parquet, so the exclusions in transfer_analysis.Rmd do
## not apply automatically and are repeated here. The two runs below sit at the
## probability floor (0.28% and 0.40% accuracy) and carry no information.
## NOTE the naming: in the parquet the reasoning-ON run carries the BARE name
## ("Qwen3.5-0.8B") and only the off-variant is suffixed. transfer_analysis.Rmd
## adds the "(reasoning on)" suffix afterwards, so a document reading the parquet
## must exclude the bare name or it silently keeps the model.
EXCLUDED_RUNS <- c("gpt-5.4-nano", "Qwen3.5-0.8B", "EuroLLM-1.7B-Instruct")

obs <- read_parquet("../../artifacts/transfer_tables/analysis.parquet") %>%
  select(source_id, language, correct, model, family, params_b, split, task,
         typological_distance, log10_common_crawl_pages) %>%
  filter(grepl("@refs/pr/16/", task, fixed = TRUE),
         !language %in% EXCLUDED_LANGUAGES,
         !model %in% EXCLUDED_RUNS)

standardise <- function(x) as.numeric((x - mean(x)) / sd(x))

cells <- obs %>%
  mutate(base_model = sub("\\s*\\(reasoning (on|off)\\)$", "", model)) %>%
  group_by(model, base_model, family, language, template = source_id, split) %>%
  summarise(n_correct = sum(correct), n_total = n(),
            params_b = first(params_b),
            typological_distance = first(typological_distance),
            log10_resource = first(log10_common_crawl_pages), .groups = "drop") %>%
  mutate(split = factor(split, levels = c("original", "synthetic")),
         log_resource_z = standardise(log10_resource),
         typological_distance_z = standardise(typological_distance),
         across(c(model, base_model, family, language, template), as.factor))
```

```{r prep-report}
cat("cells:", nrow(cells), "| languages:", nlevels(cells$language),
    "| models:", nlevels(cells$model), "\n")
cells %>% count(split, n_total)
```

# Model

Only what the claims need: the average penalty, and whether it is predicted by
the two language features. Model ability needs no fixed term --- the model,
base-model and family random effects already carry it.

```{r model-spec}
FIX <- paste("cbind(n_correct, n_total - n_correct) ~",
             "split * (log_resource_z + typological_distance_z) +")
RE_BASE <- "(1 | family/base_model) + (1 | model) + (1 | template) + (1 | model:language)"
ctrl <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 3e5))
```

```{r fit, cache=TRUE, dependson='prep', cache.extra=list(FIX, RE_BASE)}
run <- function(re_lang) {
  glmer(as.formula(paste(FIX, re_lang, "+", RE_BASE)),
        data = cells, family = binomial(), control = ctrl)
}
fit_slope <- run("(1 + split | language)")
fit_int   <- run("(1 | language)")
```

```{r convergence}
for (nm in c("fit_slope", "fit_int")) {
  f <- get(nm)
  cat(nm, "| conv:",
      if (length(f@optinfo$conv$lme4$messages)) f@optinfo$conv$lme4$messages else "OK", "\n")
}
```

# The penalty

```{r summary-slope}
summary(fit_slope)
```

# Does the penalty vary by language?

```{r lrt}
anova(fit_int, fit_slope)
vc <- VarCorr(fit_slope)$language
cat("SD of the split slope across languages:",
    round(attr(vc, "stddev")[2], 4), "log-odds\n")
```

```{r per-language}
re <- ranef(fit_slope)$language
pen <- data.frame(language = rownames(re),
                  penalty_logodds = fixef(fit_slope)[["splitsynthetic"]] + re[, 2])
pen[order(pen$penalty_logodds), ]
```
