---
title: "Cross-lingual transfer analysis"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

Which language-level features govern how well mathematical reasoning transfers
out of English? This implements the GLMM from `analysis.md`, from the raw
instance table through to the fitted model, with inspection at each stage.

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = TRUE)

library(nanoparquet)
library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)

transfer_dir <- "../../artifacts/transfer_tables"
out_dir <- "../../artifacts/analysis"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
```

# Configuration

```{r config}
SPLIT <- "synthetic" # "original" for the robustness check

# Excluded by base name (before the reasoning suffix), so each entry catches all
# of a model's reasoning variants. Drop an entry to bring a model back in.
#   gpt-5.4-nano     - proprietary: no parameter count, closed tokenizer (so no
#                      fertility). Cannot enter a model with size or fertility terms.
#   granite / Olmo-3 - excluded on request; fertility is available, so removing
#                      them here is sufficient to include them later.
EXCLUDED_MODELS <- c(
  "gpt-5.4-nano",
  "granite-3.2-2b-instruct",
  "granite-3.2-8b-instruct",
  "Olmo-3-7B-Think"
)

# eng_metric is English with metric units, not a language, and has no typological
# or resource features. uncorrected_isl is machine-translated Icelandic before
# human correction, kept in the dataset for the validation figure but not used in
# the analysis. English itself is kept -- it anchors the intercept at fertility 1,
# typological distance 0, maximum resource.
EXCLUDED_LANGUAGES <- c("eng_metric", "uncorrected_isl")
```

# Load the instance table

`collect_transfer_tables.py` writes one Parquet row per evaluated instance:
`correct` as a logical, the reasoning mode in the `model` name, and every feature
we need -- `typological_distance`, `log10_common_crawl_pages`, and the tokenizer
fertilities. Normalised fertility is the model's fertility relative to its own
English fertility, so there are no separate feature tables to join.

```{r load}
observations <- read_parquet(file.path(transfer_dir, "analysis.parquet")) %>%
  select(
    source_id, language, correct, model, family, params_b, split, eval_id,
    typological_distance, log10_common_crawl_pages,
    tokenizer_fertility, english_fertility_tokens_per_character
  ) %>%
  mutate(normalized_fertility = tokenizer_fertility / english_fertility_tokens_per_character)

dim(observations)
```

# Reasoning mode

Reasoning mode is explicit in the model name. Qwen3 / Qwen3.5 label the
reasoning-on default with a bare name and only suffix `(reasoning off)`; granite
and Olmo-3 label both sides. We give every reasoning-split model a consistent
`(reasoning on)` / `(reasoning off)` label and derive `base_model` (the tokenizer
identity, shared across modes) by stripping it.

```{r reasoning}
# Bases that appear with a "(reasoning off)" variant: their bare form is the
# reasoning-on run and should be labelled as such.
off_bases <- observations %>%
  filter(str_detect(model, "\\(reasoning off\\)$")) %>%
  mutate(base = str_remove(model, "\\s*\\(reasoning off\\)$")) %>%
  distinct(base) %>%
  pull(base)

observations <- observations %>%
  mutate(
    model = if_else(model %in% off_bases, paste0(model, " (reasoning on)"), model),
    base_model = str_remove(model, "\\s*\\(reasoning (on|off)\\)$")
  )

observations %>%
  filter(base_model %in% off_bases) %>%
  distinct(base_model, model) %>%
  arrange(base_model)
```

# Filter to the analysis set

`EXCLUDED_MODELS` is matched on `base_model`, so it catches every reasoning
variant of a listed model.

```{r filter}
observations <- observations %>%
  filter(
    split == SPLIT,
    !language %in% EXCLUDED_LANGUAGES,
    !base_model %in% EXCLUDED_MODELS
  )

count(observations, language)
count(observations, family, model)
```

# Feature coverage

Every feature is present for every included cell; the only rows the parquet leaves
without features are the ones we already drop (`gpt-5.4-nano` has no tokenizer
fertility, `eng_metric` no language features).

```{r feature-coverage}
observations %>%
  summarise(across(
    c(typological_distance, log10_common_crawl_pages, normalized_fertility),
    ~ sum(is.na(.)), .names = "na_{.col}"
  ))
```

# Aggregate to cells

One row per (model, language, template). Every predictor is constant within such
a cell, so `Binomial(n_total, pi)` has the same likelihood as the individual
Bernoulli draws at roughly 1/28th the rows -- carried through with `first()`.

```{r aggregate}
cells <- observations %>%
  group_by(model, family, language, template = source_id) %>%
  summarise(
    n_correct = sum(correct),
    n_total = n(),
    params_b = first(params_b),
    typological_distance_from_english = first(typological_distance),
    log10_common_crawl_pages = first(log10_common_crawl_pages),
    normalized_fertility = first(normalized_fertility),
    .groups = "drop"
  )

dim(cells)
```

```{r inspect-missing}
summary(cells$n_total)
```

`n_total` is 20 per cell except for `OLMo-2-0425-1B-Instruct` in `dan`, `eng` and
`nob`, where a duplicate evaluation run doubles it to 40. Those two runs are one
configuration sampled twice (near-identical accuracy, same instance ids), so
pooling them slightly overstates precision for three cells of one model; refitting
without them moved no coefficient materially. Left pooled here pending an upstream
dedup in `collect_transfer_tables.py`.

The design is otherwise balanced -- every model seen in every language, 100
templates each:

```{r inspect-balance}
table(cells$model, cells$language)
```

# Predictors

Standardise the continuous predictors so the interaction terms are interpretable
at the mean and the weakly-informative priors are on a sensible scale.

```{r standardise}
standardise <- function(x) as.numeric((x - mean(x)) / sd(x))

cells <- cells %>%
  mutate(
    log2_params = log2(params_b),
    fertility_z = standardise(normalized_fertility),
    typological_distance_z = standardise(typological_distance_from_english),
    log_resource_z = standardise(log10_common_crawl_pages),
    log_params_z = standardise(log2_params)
  ) %>%
  arrange(family, model, language, template)

cells %>%
  select(fertility_z, typological_distance_z, log_resource_z, log_params_z) %>%
  summary()
```

Two of these are language-level, so their main effects rest on 7 points
regardless of the 566k instances behind them. Worth seeing the collinearity
before reading any coefficient:

```{r inspect-collinearity}
cells %>%
  select(fertility_z, typological_distance_z, log_resource_z, log_params_z) %>%
  cor() %>%
  round(2)
```

Fertility and typological distance are strongly related, so their individual
coefficients should not be read independently.

```{r inspect-features-plot, fig.height=4}
cells %>%
  distinct(model, language, normalized_fertility, typological_distance_from_english, log10_common_crawl_pages) %>%
  pivot_longer(c(normalized_fertility, typological_distance_from_english, log10_common_crawl_pages)) %>%
  ggplot(aes(value, language)) +
  geom_point(alpha = 0.3) +
  facet_wrap(~name, scales = "free_x") +
  labs(x = NULL, y = NULL, title = "Feature spread across languages") +
  theme_minimal()
```

Raw accuracy by language and size, as a sanity check on what the model should
recover.

Languages are ordered by resource level and coloured on a sequential ramp rather
than with categorical hues: resource is a magnitude, so the colour should carry
the ordering. `mako` runs dark to light through a single blue-green hue, is
monotonic in lightness (so the order survives greyscale and colour-vision
deficiency), and avoids the pale yellow end of `viridis` that disappears against
a white panel. Highest-resource language is darkest.

```{r language-order}
language_levels <- cells %>%
  distinct(language, log10_common_crawl_pages) %>%
  arrange(desc(log10_common_crawl_pages))

# Label with the resource value so the legend states the ordering explicitly.
language_labels <- sprintf("%s (%.1f)", language_levels$language, language_levels$log10_common_crawl_pages)
names(language_labels) <- language_levels$language

language_levels
```

```{r inspect-accuracy, fig.height=4.5}
cells %>%
  group_by(model, language, log2_params) %>%
  summarise(accuracy = sum(n_correct) / sum(n_total), .groups = "drop") %>%
  mutate(language = factor(language, levels = language_levels$language)) %>%
  ggplot(aes(log2_params, accuracy, colour = language)) +
  geom_point(alpha = 0.85, size = 1.8) +
  geom_smooth(se = FALSE, method = "loess", formula = y ~ x, linewidth = 0.9) +
  scale_colour_viridis_d(
    option = "mako", begin = 0.08, end = 0.82, direction = 1,
    labels = language_labels
  ) +
  # loess is unconstrained and overshoots 1.0 at the top end; clip the view
  # rather than the data so the smoother still sees every point.
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "log2 parameters (B)", y = "accuracy",
    colour = expression(Language ~ (log[10] ~ CC ~ pages)),
    title = "Transfer gap narrows with scale",
    subtitle = "Languages ordered by resource level, darkest = highest resource"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

# Model

Language features are model-by-language (fertility) or language-only
(typological distance, resource). Interacting them with size tests the claim that
scale governs how much of the transfer gap survives.

The baseline drops the language features and keeps size; the quantity of
interest is how far `sd(model:language)` shrinks between the two, i.e. how much
of the cross-lingual transfer variation the language features explain.

Both fixed effects and both essential random terms are shared between the lme4 and
Bayesian fits; they differ only in how the nuisance groupings `template` and
`family` are handled. lme4 keeps them random (the model as written in
`analysis.md`); the Bayesian fit makes them fixed, because as random effects they
create a NUTS-hostile geometry -- the strongly-identified template SD funnels, and
the single-model family (Apertus) makes `family`/`family:model` a flat ridge that
saturates the treedepth. Fixed, they leave `sd(model:language)` unchanged. See
`transfer_spec.R` for both formulas and `transfer_brms.R` for the geometry detail.

The shared formula pieces, priors, and factor coding live in `transfer_spec.R`,
sourced here and by `transfer_brms.R`. This chunk applies the coding (sum-to-zero
contrasts, which the Bayesian fixed-effect fit uses and the lme4 random fit
ignores) and saves the prepared cells so the Bayesian job need not re-run the
786k-row prep.

```{r formulas}
source("transfer_spec.R")

cells <- code_cells_factors(cells)
saveRDS(cells, file.path(out_dir, "model_cells.rds"))
```

`script` (shared vs. not) from `analysis.md` is deliberately absent. With only
`zho` and `rus` non-Latin out of 7 languages it is nearly collinear with
typological distance and not separately identifiable -- report it descriptively
instead.

## lme4

Fast enough to iterate on, so this file keeps the frequentist fit inline; the
Bayesian version, which samples for hours, is split out into `transfer_brms.R`.
The lme4 fit keeps `template` and `family` as random effects (`LME4_RHS`) -- glmer
handles them quickly, whereas ~100 fixed template dummies make it crawl; the
Bayesian fit uses the fixed form for sampling geometry. `sd(model:language)` is
the same either way. `glmer` wants `cbind(successes, failures)` rather than brms'
`trials()` syntax.

```{r fit-lme4}
library(lme4)

fit_baseline <- glmer(as.formula(paste("cbind(n_correct, n_total - n_correct) ~", LME4_RHS_BASELINE)),
                      data = cells, family = binomial())
fit_full <- glmer(as.formula(paste("cbind(n_correct, n_total - n_correct) ~", LME4_RHS)),
                  data = cells, family = binomial())

summary(fit_full)
```

```{r inspect-lme4-variance}
VarCorr(fit_full)

sd_baseline <- attr(VarCorr(fit_baseline)$`model:language`, "stddev")[[1]]
sd_full <- attr(VarCorr(fit_full)$`model:language`, "stddev")[[1]]

tibble(
  model = c("baseline (size only)", "full (language features)"),
  sd_model_language = c(sd_baseline, sd_full),
  variance_explained = c(NA, 1 - sd_full^2 / sd_baseline^2)
)
```

## Analysis of results

KCE: the following is a work in progress

We see that size have the largest effect of parameters, with larger models performing (1.94, p<0.05) 
secondly see that higher resource languages leads to a higher performance (.699, p<0.05), but 
that the effect is less pronounced for larger models (-.210, p<0.05). 
Language with a higher typological distance from English, generally leads to worse performance (-.299, p<0.05)
and higher tokenizer fertility improved the likelihood of obtaining the correct answer (.193, p<0.05).

The interactions between tokenizer fertility and the model size (-.0576, p=.423), and
between typological distance and model size (.0332, p=.634) where not found to be significant,
indicating that effect of fertility and language distance remains a problem regardless of the size
(or at least we couldn't detect if it was influenced by it).

Beyond the individual coefficients we can ask how much of the cross-lingual transfer these
language features account for. The model-by-language random effect captures how much a model's
accuracy in a given language departs from its overall skill and the template difficulty -- the
transfer variation the fixed effects leave unexplained. Adding the three language features
shrinks this variance component from $\sigma^2 = 1.10$ (SD 1.05) in the size-only baseline to
$\sigma^2 = 0.39$ (SD 0.63) in the full model, a 64% reduction in variance (40% on the SD scale).
Roughly two-thirds of the transfer variation that remains after accounting for model skill and
template difficulty is thus captured by tokenizer fertility, typological distance and resource
level together.

Questions:
- what about resourcity and typological distance
- what about reasoning
- Is it a difference in parsing or "reasoning"
- is there more of the variance that could be explained?
- there is high correlation between fertility and language distance (estionian might fix it). 
  - One potential solution it to decompose into fertility across model (fertility between), this should be colinear 
    to typoligical distance, while the model devisation from the langauge mean fertility (fertility within) would 
    be orthogonal
- add bootstrap on proportion of variance explained (for later)

Todo:
- [ ] remove transfer spec and just add the formula in here, for the BRMS the formulas can just be incorporated into the script.
- [ ] redo the bayesian script to an Rmd. There is not need to define a function it is fine to repeat the blocks (it is two models and it reduces complexity)
- [ ] 


## Bayesian fit

The Bayesian version -- which gives posterior intervals on the variance
components, the honest form of "variance explained" given only 7 languages -- is
fit by `transfer_brms.R`, run separately because its sampling takes hours:

```
Rscript paper/scripts/analysis/transfer_brms.R
```

It reads `model_cells.rds` saved above, uses the same formula and priors from
`transfer_spec.R`, and caches its fits to `paper/artifacts/analysis/`. See that
script for the cmdstanr / dense-metric choice and the diagnostics.
