---
title: "Cross-lingual transfer -- Bayesian fit"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

The Bayesian counterpart to `transfer_analysis.Rmd`. It reads the cells that the
lme4 document prepares and saves (`model_cells.rds`) and fits the same model with
brms, which gives posterior intervals on the variance components -- the honest
form of "variance explained" given only seven languages.

Sampling takes hours, so the fits are cached to disk via brms' `file` argument: the
first knit fits and saves; later knits load. Delete the `.rds` files to refit.

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = TRUE)

library(dplyr)
library(brms)
library(cmdstanr)
library(posterior)

out_dir <- "../../artifacts/analysis"
cells <- readRDS(file.path(out_dir, "model_cells.rds"))
```

# Model form

`template` and `family` enter as fixed effects here, not as the random terms the
lme4 document uses. As random effects they give NUTS a hostile geometry -- the
strongly-identified template SD funnels, and the single-model family (Apertus)
makes `family` / `family:model` a flat ridge -- so the sampler saturates its
treedepth. Fixed, they leave `sd(model:language)` unchanged. Sum-to-zero contrasts
keep their coefficients decorrelated from the intercept.

`model` and `model:language` stay random: `model` identifies the size effect
(fixed model dummies would absorb `log_params`), and `model:language` is the
transfer variance component we are measuring.

```{r contrasts}
contrasts(cells$template) <- contr.sum(nlevels(cells$template))
contrasts(cells$family) <- contr.sum(nlevels(cells$family))
```

The backend is cmdstanr with the default diagonal metric. This posterior is
expensive for HMC -- each cell has ~2000 draws, so the likelihood pins its total
logit tightly and the additive terms that sum to it are strongly correlated, and
NUTS saturates its treedepth (a few hours per fit). It converges regardless (Rhat
~ 1, no divergences); a dense metric was tried and, at ~400 parameters, came out
worse.

Predictors are standardised, so a coefficient of 1 is a full SD of the feature
moving the log-odds by 1 -- already large. The features get a mild `normal(0, 1)`
regulariser (which stabilises the typological-distance and resource terms, resting
on 7 languages); the wide `normal(0, 5)` default covers the fixed template/family
contrasts, whose difficulty spans the logit range.

# Baseline: size only

```{r fit-baseline}
prior_baseline <-
  prior(normal(0, 2), class = "Intercept") +
  prior(normal(0, 5), class = "b") +
  prior(normal(0, 1), class = "b", coef = "log_params_z") +
  prior(student_t(3, 0, 2.5), class = "sd")

bfit_baseline <- brm(
  n_correct | trials(n_total) ~ log_params_z + template + family +
    (1 | model) + (1 | model:language),
  data = cells,
  family = binomial(),
  prior = prior_baseline,
  backend = "cmdstanr",
  chains = 4,
  iter = 2000,
  warmup = 1000,
  cores = 2,
  seed = 42,
  file = file.path(out_dir, "transfer_baseline")
)
```

# Full: language features x size

```{r fit-full}
prior_full <-
  prior(normal(0, 2), class = "Intercept") +
  prior(normal(0, 5), class = "b") +
  prior(normal(0, 1), class = "b", coef = "fertility_z") +
  prior(normal(0, 1), class = "b", coef = "typological_distance_z") +
  prior(normal(0, 1), class = "b", coef = "log_resource_z") +
  prior(normal(0, 1), class = "b", coef = "log_params_z") +
  prior(normal(0, 1), class = "b", coef = "fertility_z:log_params_z") +
  prior(normal(0, 1), class = "b", coef = "typological_distance_z:log_params_z") +
  prior(normal(0, 1), class = "b", coef = "log_resource_z:log_params_z") +
  prior(student_t(3, 0, 2.5), class = "sd")

bfit_full <- brm(
  n_correct | trials(n_total) ~
    (fertility_z + typological_distance_z + log_resource_z) * log_params_z +
    template + family + (1 | model) + (1 | model:language),
  data = cells,
  family = binomial(),
  prior = prior_full,
  backend = "cmdstanr",
  chains = 4,
  iter = 2000,
  warmup = 1000,
  cores = 2,
  seed = 42,
  file = file.path(out_dir, "transfer_full")
)

summary(bfit_full)
```

# Sampler diagnostics

Nothing below should be read until these pass: divergences zero, `Rhat` < 1.01,
and both effective sample sizes comfortably above ~400.

```{r diagnostics}
diagnostics <- function(fit) {
  np <- nuts_params(fit)
  draws <- summarise_draws(fit)
  c(
    divergent = sum(subset(np, Parameter == "divergent__")$Value),
    max_treedepth_hit = sum(subset(np, Parameter == "treedepth__")$Value >= 10),
    max_rhat = max(draws$rhat, na.rm = TRUE),
    min_bulk_ess = min(draws$ess_bulk, na.rm = TRUE),
    min_tail_ess = min(draws$ess_tail, na.rm = TRUE)
  )
}

rbind(baseline = diagnostics(bfit_baseline), full = diagnostics(bfit_full))
```

```{r trace, fig.height=7}
# Only the effects of interest; the fixed template/family contrasts would swamp it.
plot(bfit_full, variable = "^b_(Intercept|fertility|typological|log_resource|log_params)", regex = TRUE)
```

Posterior predictive check -- does the fitted model reproduce the observed counts?

```{r ppcheck, fig.height=4}
pp_check(bfit_full, ndraws = 50)
```

# Transfer variance component

The quantity of interest, as a posterior interval rather than a point estimate.
With only seven languages this interval is the honest version of "variance
explained"; the lme4 point estimate is not.

```{r variance-explained}
sd_ml_baseline <- as_draws_df(bfit_baseline)[["sd_model:language__Intercept"]]
sd_ml_full <- as_draws_df(bfit_full)[["sd_model:language__Intercept"]]

# Propagate the full posterior rather than dividing two point estimates.
reduction <- 1 - (sd_ml_full^2) / (sd_ml_baseline^2)

tibble(
  quantity = "transfer variance explained by language features",
  median = median(reduction),
  q2.5 = quantile(reduction, 0.025),
  q97.5 = quantile(reduction, 0.975)
)
```

```{r fixef}
fixef(bfit_full)
```
