---
title: "Cross-lingual transfer analysis"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

Which language-level features govern how well mathematical reasoning transfers
out of English? This implements the GLMM from `analysis.md`, from the raw
instance table through to the fitted model, with inspection at each stage.

The document fits **one** model of final accuracy and **one** decomposition of
where in a solution the transfer gap accrues. Everything else is an ablation
against the first of those.

```{r setup}
# The seven GLMM fits below take ~45 min in total, so the data-prep and fitting
# chunks are cached. Invalidation is explicit rather than trusted to knitr's
# code hashing:
#   - the prep chain carries cache.extra = md5 of the input files, so a
#     regenerated parquet forces a rebuild;
#   - the fit chunks carry cache.extra = RHS, the model formula string, so
#     editing the specification forces a refit even though the chunk code
#     itself is unchanged;
#   - dependson chains prep -> fits, so upstream changes propagate.
# cache.lazy = FALSE because the fitted merMod objects are large and lazy
# loading them is slower than reading them outright.
# Delete _cache/ to force a full rebuild.
knitr::opts_chunk$set(
  warning = FALSE, message = TRUE,
  cache.path = "_cache/transfer_analysis/",
  cache.lazy = FALSE
)

library(nanoparquet)
library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)

transfer_dir <- "../../artifacts/transfer_tables"
out_dir <- "../../artifacts/analysis"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# Hashed into the prep chunks' cache keys.
input_fingerprint <- function() {
  tools::md5sum(c(file.path(transfer_dir, "analysis.parquet"), "recovered_fertility.csv"))
}
```

# Configuration

```{r config}
SPLIT <- "synthetic" # "original" for the robustness check

# Excluded by base name (before the reasoning suffix), so each entry catches all
# of a model's reasoning variants.
#   gpt-5.4-nano - proprietary: no parameter count, closed tokenizer (so no
#                  fertility). Cannot enter a model with size or fertility terms.
EXCLUDED_MODELS <- c(
  "gpt-5.4-nano"
)

# eng_metric is English with metric units, not a language, and has no typological
# or resource features. uncorrected_isl is machine-translated Icelandic before
# human correction, kept in the dataset for the validation figure but not used in
# the analysis. English itself is kept -- it anchors the intercept at fertility 1,
# typological distance 0, maximum resource.
EXCLUDED_LANGUAGES <- c("eng_metric", "uncorrected_isl")

# Dataset revision. The parquet currently carries two:
#   refs/pr/16  - all 48 models x 16 languages, every cell n = 20. Complete and
#                 balanced.
#   d34a0ffc..  - a re-evaluation after the dan/deu/est/jpn template fixes, plus
#                 the new urd language. Partial: only 26 of 48 models for
#                 dan/deu/jpn and 17 for est.
# Pooling them would mix pre-fix and post-fix evaluations of the same templates
# (Qwen3-8B on deu scores .902 on pr16 and .920 on the pinned revision) and would
# triple-count some cells. We therefore pin to pr/16, which keeps the design
# balanced and isolates the one thing the refresh changed for us -- the number
# extraction fix, which was recomputed over every row regardless of revision.
#
# Consequences, to revisit once the re-evaluation is complete upstream:
#   - dan, deu, est and jpn use their pre-fix templates;
#   - urd is dropped (it exists only in the pinned revision, and has no
#     typological_distance or log10_common_crawl_pages in any case).
DATASET_REVISION <- "@refs/pr/16/"
```

# Load the instance table

`collect_transfer_tables.py` writes one Parquet row per evaluated instance:
`correct` as a logical, the reasoning mode in the `model` name, the number-parsing
counts, and every feature we need -- `typological_distance`,
`log10_common_crawl_pages`, and the tokenizer fertilities. Normalised fertility is
the model's fertility relative to its own English fertility.

`lhs_*` and `rhs_*` are the equation-side counts: `number_coverage_utils.py`
parses the gold solution's `<<lhs=rhs>>` calculator annotations, so `lhs_count`
is the operands a correct solution uses and `rhs_count` the intermediate results
it produces; `*_retrieved` is how many of each appear in the model's response.
These drive the reasoning-ladder section.

STOPGAP: the current parquet has **no** tokenizer fertility at all -- the column
is NA for all 48 models, where the previous version had it for all but gemma-3 and
EuroLLM-9B-Instruct. `recovered_fertility.csv` now carries the full 841
model-by-language table, lifted from the pre-refresh parquet (git 25f32c78^), and
we coalesce it in.

This is exact rather than approximate: we pin to the `refs/pr/16` revision below,
and those rows are unchanged between the two parquet versions (identical row
count and identical overall accuracy to six decimal places), so the tokenizer and
the text a fertility is computed from are both identical. The join should be
removed once the upstream fertility computation is restored.

```{r load, cache=TRUE, cache.extra=input_fingerprint()}
recovered_fertility <- read.csv("recovered_fertility.csv")

observations <- read_parquet(file.path(transfer_dir, "analysis.parquet")) %>%
  select(
    source_id, language, correct, model, family, params_b, split, eval_id, task,
    typological_distance, log10_common_crawl_pages,
    tokenizer_fertility, english_fertility_tokens_per_character,
    all_prompt_numbers_present, prompt_number_count, retrieved_prompt_number_count,
    lhs_count, lhs_retrieved, rhs_count, rhs_retrieved
  ) %>%
  left_join(recovered_fertility, by = c("model", "language"), suffix = c("", "_rec")) %>%
  mutate(
    tokenizer_fertility = coalesce(tokenizer_fertility, tokenizer_fertility_rec),
    english_fertility_tokens_per_character =
      coalesce(english_fertility_tokens_per_character, english_fertility_tokens_per_character_rec),
    relative_fertility = tokenizer_fertility / english_fertility_tokens_per_character
  ) %>%
  select(-tokenizer_fertility_rec, -english_fertility_tokens_per_character_rec)

dim(observations)
```

# Reasoning mode

Reasoning mode is explicit in the model name. Qwen3 / Qwen3.5 label the
reasoning-off runs, so the bare name is the reasoning-on run.

```{r reasoning, cache=TRUE, dependson='load'}
# Bases that appear with a "(reasoning off)" variant: their bare form is the
# reasoning-on run and should be labelled as such.
off_bases <- observations %>%
  filter(str_detect(model, "\\(reasoning off\\)$")) %>%
  mutate(base = str_remove(model, "\\s*\\(reasoning off\\)$")) %>%
  distinct(base) %>%
  pull(base)

observations <- observations %>%
  mutate(
    model = if_else(model %in% off_bases, paste0(model, " (reasoning on)"), model),
    base_model = str_remove(model, "\\s*\\(reasoning (on|off)\\)$")
  )

observations %>%
  filter(base_model %in% off_bases) %>%
  distinct(base_model, model) %>%
  arrange(base_model)
```

# Filter to the analysis set

`EXCLUDED_MODELS` is matched on `base_model`, so it catches every reasoning
variant of a listed model.

```{r filter, cache=TRUE, dependson='reasoning', cache.extra=list(SPLIT, EXCLUDED_MODELS, EXCLUDED_LANGUAGES, DATASET_REVISION)}
observations <- observations %>%
  filter(
    split == SPLIT,
    str_detect(task, fixed(DATASET_REVISION)),
    !language %in% EXCLUDED_LANGUAGES,
    !base_model %in% EXCLUDED_MODELS
  )

count(observations, language)
count(observations, family, model)

# Balance check: with a single revision pinned this must be one row per
# (model, language, template) with exactly 20 instances. If a second revision
# leaks in it shows up here as counts above 20.
observations %>%
  count(model, language, source_id, name = "n_instances") %>%
  count(n_instances)
```

# Feature coverage

```{r feature-coverage}
observations %>%
  summarise(across(
    c(typological_distance, log10_common_crawl_pages, relative_fertility),
    ~ sum(is.na(.)), .names = "na_{.col}"
  ))
```

# Aggregate to cells

One row per (model, language, template). Every predictor is constant within such
a cell, so `Binomial(n_total, pi)` has the same likelihood as the individual
Bernoulli draws at roughly 1/20th the rows -- carried through with `first()`.

```{r aggregate, cache=TRUE, dependson='filter'}
cells <- observations %>%
  group_by(model, base_model, family, language, template = source_id) %>%
  summarise(
    n_correct = sum(correct),
    n_total = n(),
    params_b = first(params_b),
    typological_distance_from_english = first(typological_distance),
    log10_common_crawl_pages = first(log10_common_crawl_pages),
    relative_fertility = first(relative_fertility),
    # Equation-side counts for the reasoning ladder.
    n_lhs = sum(lhs_count),
    n_lhs_ret = sum(lhs_retrieved),
    n_rhs = sum(rhs_count),
    n_rhs_ret = sum(rhs_retrieved),
    .groups = "drop"
  )

dim(cells)
```

```{r inspect-missing}
summary(cells$n_total)
```

`n_total` is a uniform 20 per cell -- the duplicate OLMo-2 evaluation runs that
previously doubled three cells were removed upstream in
`collect_transfer_tables.py`.

The design is balanced -- every model seen in every language, 100 templates each:

```{r inspect-balance}
table(cells$model, cells$language)
```

# Predictors

Standardise the continuous predictors so the interaction terms are interpretable
at the mean and the weakly-informative priors are on a sensible scale.

## Splitting relative fertility between and within language

Relative fertility -- a model's tokens-per-character on the target language
divided by its own tokens-per-character on English -- is the only language feature
that is a property of the *model* rather than the language, and pooling those two
roles is what made its coefficient uninterpretable in earlier versions of this
analysis. We split it:

- `relative_fertility_between` -- the language's mean relative fertility across models.
  A language-level quantity, and strongly collinear with typological distance
  (r ~ 0.8): languages that are structurally distant from English also tokenise
  badly in most models.
- `relative_fertility_within` -- a model's deviation from that language mean. This is
  "*this* model tokenises *this* language worse than its peers do", which is the
  quantity we actually have a mechanism for, and it is orthogonal by construction
  to every language-level predictor.

**Only `relative_fertility_within` enters the model.** The point of splitting was to get
the collinearity out of the fertility estimate, and `relative_fertility_between` *is* the
collinear part -- entering it puts the problem straight back in. Two reasons it
costs us nothing to drop:

- `relative_fertility_within` is unaffected. The two arms are orthogonal by construction
  (r = 0.00), so removing one leaves the other's estimate essentially unchanged.
- Typological distance is *improved*. With `relative_fertility_between` in the model, the
  typological-distance coefficient is "the effect of structural distance holding
  between-language fertility constant" -- a near-meaningless contrast when the two
  correlate at 0.8 over 16 languages, and one that inflates its standard error.
  Dropping it makes the coefficient the total language-level effect, which is both
  interpretable and better estimated.

The cost is that we can no longer claim to have tested a between-language
fertility effect separately from typological distance. We could not have done so
honestly anyway: at r = 0.8 on 16 points they are not separately identifiable, and
the null we previously reported for `relative_fertility_between` was underdetermination
rather than evidence of no effect. Typological distance should now be read as the
language-level axis those two features jointly describe -- structural distance
from English, which travels with tokenising badly.

Note this is not a return to a single pooled fertility term. The split is still
computed; we simply enter one arm of it. Pooling is what produced the wrong sign
in earlier versions.

```{r standardise, cache=TRUE, dependson='aggregate'}
standardise <- function(x) as.numeric((x - mean(x)) / sd(x))

cells <- cells %>%
  group_by(language) %>%
  mutate(relative_fertility_between = mean(relative_fertility)) %>%
  ungroup() %>%
  mutate(
    relative_fertility_within = relative_fertility - relative_fertility_between,
    log2_params = log2(params_b),
    relative_fertility_between_z = standardise(relative_fertility_between),
    relative_fertility_within_z = standardise(relative_fertility_within),
    typological_distance_z = standardise(typological_distance_from_english),
    log_resource_z = standardise(log10_common_crawl_pages),
    log_params_z = standardise(log2_params),
    # base_model is the tokenizer/weights identity shared across reasoning modes;
    # reasoning is the on/off knob (non-reasoning models are "off").
    reasoning = factor(if_else(str_detect(model, "\\(reasoning on\\)$"), "on", "off"),
                       levels = c("off", "on"))
  ) %>%
  arrange(family, model, language, template)

cells %>%
  select(relative_fertility_within_z, relative_fertility_between_z, typological_distance_z,
         log_resource_z, log_params_z) %>%
  summary()
```

Two predictors are language-level, so their main effects rest on 16 points
regardless of the 1.5M instances behind them. Worth seeing the collinearity
before reading any coefficient:

```{r inspect-collinearity}
cells %>%
  select(relative_fertility_within_z, relative_fertility_between_z, typological_distance_z,
         log_resource_z, log_params_z) %>%
  cor() %>%
  round(2)

# and at the level the language predictors actually vary
cells %>%
  distinct(language, relative_fertility_between_z, typological_distance_z, log_resource_z) %>%
  select(-language) %>%
  cor() %>%
  round(2)
```

The split does what it is meant to: `relative_fertility_within` is orthogonal (r = 0.00)
to typological distance, resource and `relative_fertility_between`, so its coefficient is
free of the collinearity problem. The r ~ 0.8 between `relative_fertility_between` and
typological distance is the reason the between arm stays out of the model -- it is
displayed here so the reader can see what is being excluded and why, not because
it is a candidate predictor.

```{r inspect-features-plot, fig.height=4}
cells %>%
  distinct(model, language, relative_fertility, typological_distance_from_english, log10_common_crawl_pages) %>%
  pivot_longer(c(relative_fertility, typological_distance_from_english, log10_common_crawl_pages)) %>%
  ggplot(aes(value, language)) +
  geom_point(alpha = 0.3) +
  facet_wrap(~name, scales = "free_x") +
  labs(x = NULL, y = NULL, title = "Feature spread across languages") +
  theme_minimal()
```

Raw accuracy by language and size, as a sanity check on what the model should
recover.

Languages are ordered by resource level and coloured on a sequential ramp rather
than with categorical hues: resource is a magnitude, so the colour should carry
the ordering. The legend states each language's absolute Common Crawl page count;
the spread is large, from ~0.6M pages for Marathi to ~880M for English. `mako` runs dark to light through a single blue-green hue, is
monotonic in lightness (so the order survives greyscale and colour-vision
deficiency), and avoids the pale yellow end of `viridis` that disappears against
a white panel. Highest-resource language is darkest.

```{r language-order}
language_levels <- cells %>%
  distinct(language, log10_common_crawl_pages) %>%
  arrange(desc(log10_common_crawl_pages))

# Label with the resource value so the legend states the ordering explicitly.
# Label with the absolute Common Crawl page count, so the legend states the
# ordering in a concrete unit rather than a log value or a ratio to English.
pages_lab <- function(l) {
  v <- 10^l
  ifelse(v >= 1e9, paste0(round(v/1e9, 1), "B"), paste0(round(v/1e6), "M"))
}
language_labels <- sprintf("%s (%s)", language_levels$language,
                           pages_lab(language_levels$log10_common_crawl_pages))
names(language_labels) <- language_levels$language

language_levels
```

```{r inspect-accuracy, fig.height=4.5}
cells %>%
  group_by(model, language, log2_params) %>%
  summarise(accuracy = sum(n_correct) / sum(n_total), .groups = "drop") %>%
  mutate(language = factor(language, levels = language_levels$language)) %>%
  ggplot(aes(log2_params, accuracy, colour = language)) +
  geom_point(alpha = 0.85, size = 1.8) +
  geom_smooth(se = FALSE, method = "loess", formula = y ~ x, linewidth = 0.9) +
  scale_colour_viridis_d(
    option = "mako", begin = 0.08, end = 0.82, direction = 1,
    labels = language_labels
  ) +
  # loess is unconstrained and overshoots 1.0 at the top end; clip the view
  # rather than the data so the smoother still sees every point.
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "log2 parameters (B)", y = "accuracy",
    colour = "Language\n(Common Crawl pages)",
    title = "Transfer gap narrows with scale",
    subtitle = "Languages ordered by resource level, darkest = highest resource"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

# The model

```{r save-cells, cache=TRUE, dependson='standardise'}
cells <- cells %>% mutate(across(c(model, base_model, family, language, template), as.factor))
saveRDS(cells, file.path(out_dir, "model_cells.rds"))
```

## Random-effects structure

Four grouping factors, each of which corresponds to a level at which one of our
predictors varies:

- `(1 | family/base_model)` -- model skill. `base_model` is the weights identity
  shared across reasoning modes.
- `(1 | model)` -- nested inside `base_model`, this is the reasoning variant. It
  matters because `reasoning` is a fixed effect that varies at exactly this
  level: without it, the reasoning main effect would be tested against cell noise
  rather than against how much reasoning helps *across* base models.
- `(1 | template)` -- item difficulty, the largest single component.
- `(1 | language)` -- the per-language residual, discussed below.
- `(1 | model:language)` -- model-specific transfer idiosyncrasy.

The `(1 | language)` term is the one worth justifying, because it is what makes
the language-level coefficients honest rather than merely large. `log_resource_z`
and `typological_distance_z` vary only across the 16 languages, but without a
language intercept their standard errors are computed from the `model:language`
stratum -- roughly (models x languages) cells treated as independent evidence
about a predictor with 16 actual points. That is pseudo-replication, and it does
not wash out with more data at the instance or model level. The `nolang`
comparison below quantifies what it costs.

The fixed-effect structure interacts the language features with size and with
reasoning, because the question is not only whether each feature hurts but
whether either of the two levers we can actually pull -- scale, reasoning --
narrows the gap it opens.

The specification lives in its own uncached chunk so that the formula strings are
always in scope, and so that the fit chunks can key their cache on `RHS`: edit the
model and every fit below rebuilds, leave it alone and none of them do.

```{r model-spec}
library(lme4)

ctrl <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 3e5))

RE <- "(1 | family/base_model) + (1 | model) + (1 | template) + (1 | language) + (1 | model:language)"
FEAT <- "(relative_fertility_within_z + typological_distance_z + log_resource_z)"
RHS <- paste(
  FEAT, "* log_params_z +",
  "reasoning * (relative_fertility_within_z +",
  "typological_distance_z + log_resource_z + log_params_z) +", RE
)

cat(RHS)
```

```{r fit-lme4, cache=TRUE, dependson='save-cells', cache.extra=RHS}
fit_main <- glmer(
  as.formula(paste("cbind(n_correct, n_total - n_correct) ~", RHS)),
  data = cells, family = binomial(), control = ctrl
)
```

```{r summary-main}
summary(fit_main)
```

The baseline drops the language features and keeps size, with the same random
structure. Comparing variance components between the two is how we read off the
transfer variation the features explain.

```{r fit-baseline, cache=TRUE, dependson='save-cells', cache.extra=RE}
fit_baseline <- glmer(
  as.formula(paste("cbind(n_correct, n_total - n_correct) ~ log_params_z +", RE)),
  data = cells, family = binomial(), control = ctrl
)
```

```{r summary-baseline}
summary(fit_baseline)

VarCorr(fit_main)
VarCorr(fit_baseline)
```

## Results

Size is the largest single effect (`log_params_z` ~ 1.88, p < 1e-16). Among the
language features, resource level dominates (`log_resource_z` ~ 0.71, p < 1e-16):
higher-resource languages are substantially easier. Typological distance hurts
(~ -0.29, p = 1.5e-4), and within-language relative fertility hurts in the theoretically
expected direction (~ -0.07, p = .019) -- a model that tokenises a language worse
than its peers do performs worse on it.

Read the typological-distance coefficient as the whole language-level axis of
structural distance from English, which travels with tokenising that language
badly (r ~ 0.8 between the two over our 16 languages). It is not a distance effect
net of tokenisation; the design cannot separate those.

That sign on fertility is worth flagging: pooled fertility carried a *positive*
coefficient in earlier versions of this analysis, which never made sense. It was
an artifact of the between-language collinearity, and splitting the term both
fixes the sign and explains the old result.

The interactions are where the actionable content sits, and the two levers behave
differently:

- **Scale buffers resource.** `log_resource_z:log_params_z` ~ -0.25, p < 1e-16.
  Larger models suffer less from a language being low-resource.
- **Scale does not buffer typological distance.** ~ 0.02, p = .45. Nor fertility
  (p = .42).
- **Reasoning buffers resource.** `log_resource_z:reasoningon` ~ -0.15, p = .004.
- **Reasoning also buffers typological distance.** ~ +0.12, p = .013. The main
  effect is negative, so a positive interaction means the distance penalty shrinks
  when reasoning is on.

So scale helps only against data scarcity, while reasoning helps against both.
Reasoning is the only lever here that does anything about structural distance --
which is the more useful finding of the two, because distance is the feature you
cannot change about a language.

A caveat on that last bullet, because it is new. In an earlier specification that
also carried a between-language fertility term, this interaction was ~0.03 with
p = .67, and the document concluded that neither lever touched structural
distance. The two terms are collinear at r ~ 0.8 and were splitting one effect
between them (`relative_fertility_between:reasoning` took ~0.12, p = .13). Dropping the
between arm assigns it to typological distance, which is the interpretation the
design supports -- but it means the effect is identified off a language-level axis
that confounds structural distance with tokenisation quality, and should be
described that way rather than as "reasoning helps with grammar".

### Effect sizes in interpretable units

Coefficients are per standard deviation of a standardised predictor, which is
convenient for fitting and useless for intuition. Both size and resource are
log-transformed, so dividing by the SD puts them in *doublings*, and the two can
then be quoted against each other.

Note that standardising makes the choice of log base irrelevant to the fit --
`z(log10 x)` and `z(log2 x)` are the same vector to numerical precision -- so this
is purely a reporting transformation and changes nothing estimated above.

```{r effect-units}
sd_p <- sd(cells$log2_params)
sd_r <- sd(cells$log10_common_crawl_pages)
b <- fixef(fit_main)

per_doubling_params <- b[["log_params_z"]] / sd_p
per_doubling_pages  <- b[["log_resource_z"]] / sd_r * log10(2)

tibble::tibble(
  quantity = c("per doubling of parameters", "per doubling of Common Crawl pages",
               "doublings of data equal to one doubling of parameters"),
  value = c(per_doubling_params, per_doubling_pages,
            per_doubling_params / per_doubling_pages)
)
```

Model size buys about **1.0 log-odds per doubling of parameters**, and language
resource about **0.25 log-odds per doubling of Common Crawl pages**. So one
doubling of model size is worth roughly **four doublings of language data**, i.e.
about 15x the data -- a convenient way to state the trade-off, and directly
comparable to how scaling laws are usually reported.

The same exchange rate prices the language gaps themselves: the observed range of
resource levels, English down to Marathi, costs about **2.7 doublings of model
size (~6x the parameters)**, and the range of typological distance about **1.1
doublings (~2x)**.

Two caveats on these numbers. They use main effects at the sample means and
therefore ignore the `resource x size` interaction, which makes the true exchange
rate depend on model size; and Common Crawl pages is a proxy for a language's
share of the training mixture, not a measurement of it.

The reasoning **main** effect is not significant (~ 0.49, p = .12). This is a
consequence of the `(1 | model)` term putting it on the correct stratum: it is
tested against the 12 base models run both ways, which is genuinely thin evidence
for an average lift. The claim this model supports is the interaction -- reasoning
differentially helps languages that are far from English on both axes -- not
"reasoning raises accuracy".
For the latter, the paired comparison further down is better evidence and does
not depend on this specification at all.

## How much transfer variation do the features explain?

With a language intercept present, the question splits into two, and they have
very different answers.

```{r variance-explained}
vc <- function(f, g) attr(VarCorr(f)[[g]], "stddev")[[1]]

tibble(
  component = c("language", "model:language"),
  sd_baseline = c(vc(fit_baseline, "language"), vc(fit_baseline, "model:language")),
  sd_full = c(vc(fit_main, "language"), vc(fit_main, "model:language"))
) %>%
  mutate(variance_explained = 1 - sd_full^2 / sd_baseline^2)
```

- **Between-language variance: ~89% explained** (SD 0.83 -> 0.27). Almost all of
  *why some languages are harder than others* is captured by resource level,
  typological distance and fertility.
- **Model-by-language variance: ~20% explained** (SD 0.65 -> 0.58). *Which* model
  struggles with *which* language, beyond that language's general difficulty, is
  largely unexplained. `relative_fertility_within` is the only predictor operating at this
  level and its effect is small.

The second number is a real negative result and belongs in the write-up: language
features are close to sufficient for language difficulty, and close to useless for
model-specific transfer failure.

Note that earlier versions of this analysis reported a single pooled figure
(~64% of `sd(model:language)`) because there was no language intercept to
separate the two components; that number is not comparable to either of these.

## Which languages do we predict poorly?

```{r language-residuals}
ranef(fit_main)$language %>%
  tibble::rownames_to_column("language") %>%
  arrange(`(Intercept)`)
```

The features over-predict `rus`, `deu`, `isl` and `est`, and under-predict `nob`,
`ara`, `ita` and `dan`. Russian is the striking case: second-highest resource and
low typological distance, so the model expects a great deal and does not get it.

There is a candidate explanation worth one follow-up. The languages that
*over*-perform their predictions -- Norwegian, Danish, Italian -- all have close,
high-resource relatives (Swedish/German/English; French/Spanish), while our
features only measure distance to *English*. A resource-weighted typological
proximity to the whole training mix, rather than to English alone, would test
that. Treat it as a hypothesis: 16 shrunken residuals will support many stories.

`zho` at +0.14 is consistent with Common Crawl page counts underestimating
Chinese resource, which is a known weakness of the proxy.

## What the language intercept costs

Refitting the identical model without `(1 | language)` isolates the effect of
using the wrong error stratum.

```{r fit-nolang, cache=TRUE, dependson='save-cells', cache.extra=RHS}
fit_nolang <- glmer(
  as.formula(paste("cbind(n_correct, n_total - n_correct) ~",
                   sub("\\(1 \\| language\\) \\+ ", "", RHS))),
  data = cells, family = binomial(), control = ctrl
)
```

```{r nolang-comparison}
summary(fit_nolang)

tm <- summary(fit_main)$coefficients
tn <- summary(fit_nolang)$coefficients
keep <- intersect(rownames(tm), rownames(tn))
round(data.frame(
  est_nolang = tn[keep, 1], se_nolang = tn[keep, 2], p_nolang = tn[keep, 4],
  est_main = tm[keep, 1], se_main = tm[keep, 2], p_main = tm[keep, 4],
  se_ratio = tm[keep, 2] / tn[keep, 2]
), 4)
```

The result is about as clean as this kind of check gets: standard errors inflate
by ~2.6x on *exactly* the two language-level predictors and are unchanged
(0.90-1.00x) on every other term. Point estimates barely move (`log_resource_z`
0.7092 -> 0.7119, `typological_distance_z` -0.2882 -> -0.2876), so this is purely
an inference correction, not a different answer.

The consequential case is typological distance: z = -9.8 without the intercept,
-3.8 with it. Both are significant, but the first overstates the evidence by a
factor the design does not support. Resource behaves the same way (z = 24.1 to
9.4) and survives comfortably either way.

# Ablations

Two additions the earlier analysis flagged as promising. Both are null once
tested on the correct stratum with 16 languages.

## Resource x typological distance

Does a high resource level buffer the typological-distance penalty?

```{r ablation-resource-typology, cache=TRUE, dependson=c('save-cells','fit-lme4'), cache.extra=RHS}
fit_rxt <- update(fit_main, . ~ . + typological_distance_z:log_resource_z)
```

```{r ablation-resource-typology-result}
round(summary(fit_rxt)$coefficients["typological_distance_z:log_resource_z", , drop = FALSE], 4)
anova(fit_main, fit_rxt)

summary(fit_rxt)
```

No: ~0.02, p = .68. Earlier versions found this interaction significant, but that
was a third language-level term tested against `model:language` variance -- the
same stratum error the language intercept fixes. It is the sharpest illustration
in this document of why the random structure matters.

## Script

`script` was previously dropped as unidentifiable. With 7 non-Latin languages out
of 16 it now *is* identifiable, though the leverage is thin: it comes almost
entirely from Cyrillic `rus`/`ukr` (non-Latin but typologically close to English)
against Latin-script `est` (typologically distant).

```{r ablation-script, cache=TRUE, dependson='save-cells', cache.extra=RHS}
non_latin <- c("ara", "hin", "jpn", "mar", "rus", "ukr", "urd", "zho")
cells_script <- cells %>%
  mutate(non_latin = factor(if_else(language %in% non_latin, "yes", "no"), levels = c("no", "yes")))

fit_script <- glmer(as.formula(paste("cbind(n_correct, n_total - n_correct) ~ non_latin +", RHS)),
                    data = cells_script, family = binomial(), control = ctrl)
```

```{r ablation-script-result}
round(summary(fit_script)$coefficients[c("non_latinyes", "typological_distance_z"), , drop = FALSE], 4)

summary(fit_script)
```

Also null (~ -0.03, p = .90). Note the converse is informative too: typological
distance barely moves when script is added (-0.29 -> -0.28, still p < .05), so the
typological-distance effect is not a writing-system effect in disguise. The URIEL
distance we use is cosine distance from English over the 103-feature `syntax_knn`
vector -- purely morphosyntactic (word order, adposition and adjective order, case
marking), a property of the spoken language -- so it carries no orthographic
information to begin with.

Note that `transferfeatures.py` describes this feature in its module docstring and
in the exported `feature_sources.json` as concatenating `syntax_knn` *and*
`inventory_knn`. The code does not: `collect_language_features()` requests
`syntax_knn` alone, and records the accurate label in `typological_feature_set`.
The prose provenance is stale and should be corrected upstream.

# Reasoning and the transfer gap

Because reasoning is a knob rather than a property of a language, its average
effect is clearest from the paired models directly -- reasoning on versus off,
across the base models run both ways.

```{r reasoning-transfer-gap}
paired_bases <- cells %>%
  distinct(base_model, reasoning) %>%
  count(base_model) %>%
  filter(n == 2) %>%
  pull(base_model)

by_lang <- cells %>%
  filter(base_model %in% paired_bases) %>%
  group_by(reasoning, language) %>%
  summarise(acc = sum(n_correct) / sum(n_total), .groups = "drop") %>%
  mutate(logit = qlogis(acc))

english <- by_lang %>% filter(language == "eng") %>% select(reasoning, eng_acc = acc, eng_logit = logit)

by_lang %>%
  filter(language != "eng") %>%
  left_join(english, by = "reasoning") %>%
  group_by(reasoning) %>%
  summarise(
    mean_accuracy       = mean(acc),
    gap_accuracy_points = mean(eng_acc - acc),      # absolute: English minus target
    # Two relative measures, because they are NOT the same and the difference
    # is large at our accuracy levels. The odds ratio is what the GLMM is linear
    # in; the risk ratio is what the phrase "times more likely to be wrong"
    # actually means. Reporting the former with the latter's wording overstates
    # the gap by 40-85% here.
    odds_ratio          = mean(exp(eng_logit - logit)),
    risk_ratio          = mean((1 - acc) / (1 - eng_acc))
  )
```

Turning reasoning on raises accuracy across the board (0.62 to 0.69, ~+7 points)
and shrinks the English-to-target gap from 20.6 to 12.9 accuracy points. It also
shrinks on the *relative* scale, on both of the measures below: with reasoning
off a model is on average **2.2x more likely** to answer incorrectly in a
non-English language than in English, falling to 1.7x with reasoning on; on the
odds scale the same comparison is 3.1x falling to 2.1x.

Keep those two apart. "Times more likely to be wrong" is a **risk** ratio; the
GLMM is linear in the **odds** ratio, and at our accuracy levels they diverge
badly -- 3.1 versus 2.2 on average, and 6.0 versus 3.2 for Marathi. Quoting an
odds ratio with risk-ratio wording is the standard misreading and inflates the
claim substantially, so state which one is meant. We report the risk ratio, as
the quantity the plain-language phrasing actually names.

One consequence to keep in view when reading the figures against the coefficient
table. A risk ratio depends on the base rate it is measured against: at 32B the
English error rate is ~7%, so a given absolute drop in accuracy produces a larger
ratio than the same drop at 4B, where the English error rate is far higher. On the
risk-ratio scale the large model therefore shows the *steeper* resource slope,
while the fitted interaction -- estimated on the odds scale -- is
`log_resource_z:log_params_z` = -0.25, meaning scale *reduces* the resource
penalty. Both are correct statements about different quantities. The inferential
claim is the coefficient; the risk-ratio figure is the descriptive one.

The distinction between the two gap measures matters. A uniform lift narrows the
absolute gap on its own, because the accuracy scale compresses near the ceiling,
while leaving the error-odds ratio unchanged. The ratio falling from 3.1 to 2.1
means reasoning does more than raise all boats -- and the fitted term behind that
is the `reasoning x resource` interaction in the main model (~ -0.13, p = .009).

This paired comparison, not the model's reasoning main effect, is the right
evidence for "reasoning helps". The two answer different questions and both
belong in the paper.

# Where in the solution does the gap accrue?

The earlier version of this section asked a binary question -- is the gap about
parsing the problem or doing the maths? -- using whether the response reproduced
the prompt's numbers. The equation-side counts let us ask a graded version
instead, which is both sharper and avoids a serious measurement problem with the
prompt-number metric (see the caveat at the end of this section).

Three rungs, increasing in depth through the solution:

1. **LHS** -- the operands a correct solution uses. Roughly "did the model take
   in the right inputs?"
2. **RHS** -- the intermediate results a correct solution produces. "Did the
   model actually compute the intervening steps?"
3. **Final answer** -- `correct`, i.e. the main model above.

All three are marginal, not conditional: rung 2 does not condition on having
passed rung 1. So the ladder measures *total accumulated* degradation at
increasing depth, which is what we want for a claim about compounding, and it
avoids conditioning on a post-treatment variable.

```{r ladder-descriptive}
cells %>%
  group_by(language) %>%
  summarise(
    lhs_rate = sum(n_lhs_ret) / sum(n_lhs),
    rhs_rate = sum(n_rhs_ret) / sum(n_rhs),
    accuracy = sum(n_correct) / sum(n_total),
    .groups = "drop"
  ) %>%
  arrange(desc(accuracy))
```

```{r ladder-validation}
# The chevron denominators come from the gold solution, so they should be
# language-invariant: same templates, same equations. If they are not, the
# rungs are not comparable across languages.
cells %>%
  group_by(language) %>%
  summarise(mean_lhs = sum(n_lhs) / sum(n_total),
            mean_rhs = sum(n_rhs) / sum(n_total),
            .groups = "drop") %>%
  arrange(mean_lhs)

# cells with no chevrons at all, which drop out of the ladder fits
sum(cells$n_lhs == 0)
```

`lhs_count` and `rhs_count` are near-identical across all 16 languages and only
~2% of cells have no chevrons, uniformly distributed, so the denominators are
comparable and the rungs measure what we intend.

```{r fit-ladder, cache=TRUE, dependson='save-cells', cache.extra=RHS}
fit_lhs <- glmer(as.formula(paste("cbind(n_lhs_ret, n_lhs - n_lhs_ret) ~", RHS)),
                 data = filter(cells, n_lhs > 0), family = binomial(), control = ctrl)

fit_rhs <- glmer(as.formula(paste("cbind(n_rhs_ret, n_rhs - n_rhs_ret) ~", RHS)),
                 data = filter(cells, n_rhs > 0), family = binomial(), control = ctrl)
```

The comparison that matters is across rungs, so the side-by-side comes first; the
two fits in full follow.

```{r ladder-comparison}
ladder_terms <- c("relative_fertility_within_z", "typological_distance_z",
                  "log_resource_z", "log_params_z", "log_resource_z:log_params_z")
round(data.frame(
  lhs   = summary(fit_lhs)$coefficients[ladder_terms, 1],
  rhs   = summary(fit_rhs)$coefficients[ladder_terms, 1],
  final = summary(fit_main)$coefficients[ladder_terms, 1]
), 4)
```

```{r summary-ladder}
summary(fit_lhs)
summary(fit_rhs)
```

## Reading the ladder

Every coefficient grows along the chain: typological distance
-0.09 -> -0.17 -> -0.29 (3.2x end to end), resource 0.30 -> 0.42 -> 0.71 (2.4x),
within-language relative fertility -0.03 -> -0.07 -> -0.07 (2.0x). Size grows fastest
(0.49 -> 0.93 -> 1.88, 3.8x), as does the size-buffers-resource interaction
(-0.07 -> -0.12 -> -0.25).

**The marginal ladder cannot, on its own, establish that this is compounding.**
The three outcomes have different base rates and different residual dispersion, so
logit coefficients are not directly comparable in magnitude across them. A pure
scale difference between the outcomes would inflate every coefficient by roughly a
common factor -- which is very close to what we observe. There is no feature here
that stays flat and could serve as an internal control.

An earlier version of this analysis did report such a control: typological
distance appeared flat across the rungs (-0.28, -0.27, -0.29) while resource
tripled, and the contrast was read as evidence that structural distance is a cost
paid once at input while data scarcity accumulates. That pattern was an artifact
of a number-extraction bug. Japanese is the most typologically distant language in
the set, and unmatched kanji numerals artificially depressed its LHS and RHS rates
-- the two rungs that depend on extracting numbers from responses -- which
inflated the typological-distance coefficient at exactly those rungs. With the
extraction fixed, Japanese moves from 0.68/0.59 to 0.78/0.64 and the coefficient
at the LHS rung drops by two thirds. The conclusion did not survive the fix, and
the "fixed tax versus accumulating penalty" reading should not be quoted from
earlier drafts -- typological distance is now the *fastest*-growing of the three,
the opposite of what that reading claimed.

So read the ladder as descriptive: the gap widens with depth on the raw rates
(English to Marathi runs 0.19 at LHS, 0.30 at RHS, 0.36 at the final answer), and
leave the question of *why* to the adjusted analysis below, which compares
coefficients within a single outcome and is not exposed to the scale objection.

`sd(language)` also grows along the ladder (0.14 -> 0.16 -> 0.27), and the
`family` variance collapses to the boundary at the LHS rung -- model family is
irrelevant to retrieving numbers and dominant for arithmetic. That last contrast
is on the same scale caveat, but a variance component going to zero is not a
matter of degree.

```{r ladder-variance}
sdl <- function(f, g) attr(VarCorr(f)[[g]], "stddev")[[1]]
tibble(
  rung = c("lhs", "rhs", "final"),
  language = c(sdl(fit_lhs, "language"), sdl(fit_rhs, "language"), sdl(fit_main, "language")),
  model_language = c(sdl(fit_lhs, "model:language"), sdl(fit_rhs, "model:language"), sdl(fit_main, "model:language")),
  family = c(sdl(fit_lhs, "family"), sdl(fit_rhs, "family"), sdl(fit_main, "family"))
)
```

## Per-step versus accumulated

The marginal ladder cannot separate two mechanisms: each reasoning step being
individually harder in low-resource languages, versus each step being equally hard
with the deficit simply accumulating over more steps.

Conditioning hard on a full operand parse does not work as a test. Among
instances where every operand was retrieved, the RHS rate is ~97% -- at ceiling,
with nothing left to explain -- and the conditioning is severely differential
across languages (full-LHS rates run from ~0.34 for `mar` to ~0.61 for `eng`), so
it would compare the top third of one language against the top two thirds of
another. That is selection on a collider, and it would produce a table that looks
meaningful and is not.

Adjusting for operand retrieval instead of conditioning on it keeps the full
sample and avoids the ceiling. It is still mediator adjustment, so the coefficient
is not a clean causal effect, but it is statistically well behaved.

```{r fit-rhs-adjusted, cache=TRUE, dependson='save-cells', cache.extra=RHS}
cells_adj <- cells %>%
  filter(n_lhs > 0, n_rhs > 0) %>%
  mutate(lhs_rate_z = as.numeric(scale(n_lhs_ret / n_lhs)))

fit_rhs_adj <- glmer(
  as.formula(paste("cbind(n_rhs_ret, n_rhs - n_rhs_ret) ~ lhs_rate_z +", RHS)),
  data = cells_adj, family = binomial(), control = ctrl
)
```

```{r rhs-adjusted-comparison}
tt <- c("log_resource_z", "typological_distance_z", "relative_fertility_within_z")
round(data.frame(
  unadjusted = summary(fit_rhs)$coefficients[tt, 1],
  adjusted = summary(fit_rhs_adj)$coefficients[tt, 1],
  adj_se = summary(fit_rhs_adj)$coefficients[tt, 2],
  adj_p = summary(fit_rhs_adj)$coefficients[tt, 4]
), 4)

summary(fit_rhs_adj)
```

All three language features behave the same way: most of their effect on producing
intermediate results is inherited from how well the operands were read, and a
small but statistically clear residual survives.

| feature | unadjusted | adjusted | share propagated | adjusted p |
|---|---|---|---|---|
| `log_resource_z` | 0.420 | 0.099 | 76% | < 1e-4 |
| `typological_distance_z` | -0.172 | -0.075 | 56% | < 1e-4 |
| `relative_fertility_within_z` | -0.067 | -0.033 | 50% | .005 |

Because this comparison is made *within a single outcome*, it is not exposed to
the cross-outcome scale objection that limits the marginal ladder, which is why it
carries the interpretation:

> **The transfer gap is predominantly a failure to take in the problem correctly.
> Once a model has read the operands, its remaining per-step disadvantage is real
> but small -- roughly a quarter to a half of the raw effect.**

This is consistent with the conditional descriptive above: among instances where
every operand was retrieved, intermediate results follow ~97% of the time
regardless of language. The arithmetic itself is close to language-invariant; what
differs is getting the right numbers into it.

The three features differ in degree rather than kind, and the ordering is
interpretable: resource is the most propagated (76%), typological distance less so
(56%), fertility least (50%). Data scarcity does most of its damage by the model
failing to take the problem in; structural distance and bad tokenisation keep more
of a per-step penalty. The gap between 56% and 50% rests on the two smallest
coefficients in the model, so treat that end of the ordering as suggestive.

Two caveats, and the second cuts against the residuals:

1. This is mediator adjustment, not identified mediation. Anything affecting both
   operand retrieval and intermediate production without being in the model --
   item-specific model competence, say -- can bias the adjusted coefficients. The
   random effects absorb some of this but not all.
2. `lhs_rate_z` measures operand retrieval imperfectly, and adjusting for a noisily
   measured mediator *under*-adjusts. The residual direct effects are therefore
   biased away from zero, so the adjusted column is best read as an upper bound on
   the true per-step penalties. The propagated shares are correspondingly lower
   bounds -- the reading-stage story is if anything stronger than the table shows.

## Caveat: the prompt-number metric is not usable

The previous version of this section used `all_prompt_numbers_present`. It is
still not usable for Japanese, though the failure mode has changed.

The prompt numbers are recovered by extracting from the prompt text, so the metric
is only comparable across languages if extraction is equally reliable in each. It
was not. In the earlier parquet, Japanese kanji numerals went unmatched and ~93%
of Japanese prompts yielded zero extractable numbers (mean 0.08 against 3.24 for
English); because an empty set evaluated vacuously true, Japanese was recorded as
the *best*-parsing language in the study, above English.

The upstream refresh fixed the under-extraction, and fixed it broadly: `ara`
(2.74 -> 3.22), `nld` (3.08 -> 3.23), `ita` (3.07 -> 3.22) and `zho` (3.13 -> 3.26)
all moved into line, and fifteen of sixteen languages now sit within 0.13 numbers
per template of English.

Japanese overshot. It now extracts **4.37** numbers per prompt against English's
3.24 -- an excess of 1.13 per template, on 82 of 100 templates -- with
`retrieved_prompt_number_count` inflated to match (3.91 against 3.10). The
templates are identical across languages, so this is spurious matching, and kanji
numerals inside ordinary words are the obvious candidate: 一緒, 一番, 十分 and 二人
all contain characters that are numerals standing alone. The bias has flipped
sign rather than disappeared -- more spurious prompt numbers makes "all present"
harder to satisfy, so Japanese should now look artificially *bad* where it
previously looked artificially good.

```{r prompt-number-check}
# The templates are shared, so prompt_number_count should be language-invariant.
# Any language far from the others is an extraction artifact, not a finding.
cells_pn <- observations %>%
  group_by(language) %>%
  summarise(mean_prompt_numbers = mean(prompt_number_count),
            mean_retrieved = mean(retrieved_prompt_number_count),
            .groups = "drop") %>%
  arrange(desc(mean_prompt_numbers))
cells_pn
```

The ladder is unaffected by any of this, which is why it is the instrument this
document uses. Its denominators come from the gold solution's `<<lhs=rhs>>`
chevrons, which are written in ASCII digits in every language -- verified
language-invariant in the validation chunk above -- and its numerators from model
responses, which use ASCII digits too. Japanese sits mid-table on both rungs, in
line with its accuracy.

# Notes and limitations

Our resource proxy is Common Crawl page count. While probably close to real-world
distributions, it likely has systematic bias (Chinese looks underestimated, and
its positive language residual is consistent with that) and does not vary across
model. Ideally this analysis would use models whose training distribution is
known.

The reasoning main effect is identified from the 12 base models run both ways;
Olmo-3-7B-Think is reasoning-on only and contributes only between-base
information. Worth confirming the coefficient is stable on paired bases alone.

Outstanding:

- [ ] bootstrap the variance-explained proportions (`bootMer`)
- [ ] resource-weighted typological proximity to the training mix, rather than
      distance to English alone -- suggested by the language residuals
- [ ] confirm the reasoning main effect on paired bases only
- [ ] upstream: kanji numeral extraction in `number_coverage_utils.py`
