---
title: "Cross-lingual transfer analysis"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

Which language-level features govern how well mathematical reasoning transfers
out of English? This implements the GLMM from `analysis.md`, from the raw
instance table through to the fitted model, with inspection at each stage.

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = TRUE)

library(nanoparquet)
library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)

transfer_dir <- "../../artifacts/transfer_tables"
out_dir <- "../../artifacts/analysis"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
```

# Configuration

```{r config}
SPLIT <- "synthetic" # "original" for the robustness check

# Excluded by base name (before the reasoning suffix), so each entry catches all
# of a model's reasoning variants.
#   gpt-5.4-nano - proprietary: no parameter count, closed tokenizer (so no
#                  fertility). Cannot enter a model with size or fertility terms.
# Granite and Olmo-3 (excluded earlier) are back in: they are reasoning-capable,
# which the reasoning section below needs. Granite adds a non-Qwen on/off pair;
# Olmo-3-7B-Think is reasoning-on only in the synthetic split.
EXCLUDED_MODELS <- c(
  "gpt-5.4-nano"
)

# eng_metric is English with metric units, not a language, and has no typological
# or resource features. uncorrected_isl is machine-translated Icelandic before
# human correction, kept in the dataset for the validation figure but not used in
# the analysis. English itself is kept -- it anchors the intercept at fertility 1,
# typological distance 0, maximum resource.
EXCLUDED_LANGUAGES <- c("eng_metric", "uncorrected_isl")
```

# Load the instance table

`collect_transfer_tables.py` writes one Parquet row per evaluated instance:
`correct` as a logical, the reasoning mode in the `model` name, the number-parsing
counts, and every feature we need -- `typological_distance`,
`log10_common_crawl_pages`, and the tokenizer fertilities. Normalised fertility is
the model's fertility relative to its own English fertility.

STOPGAP: the regenerated parquet dropped tokenizer fertility for gemma-3 (all
sizes) and EuroLLM-9B-Instruct. `recovered_fertility.csv` holds those values, taken
from the previous parquet (git 19edd05); we coalesce them in until the upstream
fertility computation is fixed, after which this join can be removed.

```{r load}
recovered_fertility <- read.csv("recovered_fertility.csv")

observations <- read_parquet(file.path(transfer_dir, "analysis.parquet")) %>%
  select(
    source_id, language, correct, model, family, params_b, split, eval_id,
    typological_distance, log10_common_crawl_pages,
    tokenizer_fertility, english_fertility_tokens_per_character,
    all_prompt_numbers_present, prompt_number_count, retrieved_prompt_number_count
  ) %>%
  left_join(recovered_fertility, by = c("model", "language"), suffix = c("", "_rec")) %>%
  mutate(
    tokenizer_fertility = coalesce(tokenizer_fertility, tokenizer_fertility_rec),
    english_fertility_tokens_per_character =
      coalesce(english_fertility_tokens_per_character, english_fertility_tokens_per_character_rec),
    normalized_fertility = tokenizer_fertility / english_fertility_tokens_per_character
  ) %>%
  select(-tokenizer_fertility_rec, -english_fertility_tokens_per_character_rec)

dim(observations)
```

# Reasoning mode

Reasoning mode is explicit in the model name. Qwen3 / Qwen3.5 label the
reasoning-on default with a bare name and only suffix `(reasoning off)`; granite
and Olmo-3 label both sides. We give every reasoning-split model a consistent
`(reasoning on)` / `(reasoning off)` label and derive `base_model` (the tokenizer
identity, shared across modes) by stripping it.

```{r reasoning}
# Bases that appear with a "(reasoning off)" variant: their bare form is the
# reasoning-on run and should be labelled as such.
off_bases <- observations %>%
  filter(str_detect(model, "\\(reasoning off\\)$")) %>%
  mutate(base = str_remove(model, "\\s*\\(reasoning off\\)$")) %>%
  distinct(base) %>%
  pull(base)

observations <- observations %>%
  mutate(
    model = if_else(model %in% off_bases, paste0(model, " (reasoning on)"), model),
    base_model = str_remove(model, "\\s*\\(reasoning (on|off)\\)$")
  )

observations %>%
  filter(base_model %in% off_bases) %>%
  distinct(base_model, model) %>%
  arrange(base_model)
```

# Filter to the analysis set

`EXCLUDED_MODELS` is matched on `base_model`, so it catches every reasoning
variant of a listed model.

```{r filter}
observations <- observations %>%
  filter(
    split == SPLIT,
    !language %in% EXCLUDED_LANGUAGES,
    !base_model %in% EXCLUDED_MODELS
  )

count(observations, language)
count(observations, family, model)
```

# Feature coverage

Every feature is present for every included cell; the only rows the parquet leaves
without features are the ones we already drop (`gpt-5.4-nano` has no tokenizer
fertility, `eng_metric` no language features).

```{r feature-coverage}
observations %>%
  summarise(across(
    c(typological_distance, log10_common_crawl_pages, normalized_fertility),
    ~ sum(is.na(.)), .names = "na_{.col}"
  ))
```

# Aggregate to cells

One row per (model, language, template). Every predictor is constant within such
a cell, so `Binomial(n_total, pi)` has the same likelihood as the individual
Bernoulli draws at roughly 1/28th the rows -- carried through with `first()`.

```{r aggregate}
cells <- observations %>%
  group_by(model, family, language, template = source_id) %>%
  summarise(
    n_correct = sum(correct),
    n_total = n(),
    params_b = first(params_b),
    typological_distance_from_english = first(typological_distance),
    log10_common_crawl_pages = first(log10_common_crawl_pages),
    normalized_fertility = first(normalized_fertility),
    # Number-parsing counts, summed over the cell's instances, for the
    # "parsing vs reasoning" section below.
    n_retrieved = sum(retrieved_prompt_number_count),   # prompt numbers found in the response
    n_prompt_numbers = sum(prompt_number_count),        # prompt numbers to find
    n_parsed = sum(all_prompt_numbers_present),          # instances with all numbers found
    n_correct_parsed = sum(correct & all_prompt_numbers_present),
    .groups = "drop"
  )

dim(cells)
```

```{r inspect-missing}
summary(cells$n_total)
```

`n_total` is 20 per cell except for `OLMo-2-0425-1B-Instruct` in `dan`, `eng` and
`nob`, where a duplicate evaluation run doubles it to 40. Those two runs are one
configuration sampled twice (near-identical accuracy, same instance ids), so
pooling them slightly overstates precision for three cells of one model; refitting
without them moved no coefficient materially. Left pooled here pending an upstream
dedup in `collect_transfer_tables.py`.

The design is otherwise balanced -- every model seen in every language, 100
templates each:

```{r inspect-balance}
table(cells$model, cells$language)
```

# Predictors

Standardise the continuous predictors so the interaction terms are interpretable
at the mean and the weakly-informative priors are on a sensible scale.

```{r standardise}
standardise <- function(x) as.numeric((x - mean(x)) / sd(x))

cells <- cells %>%
  mutate(
    log2_params = log2(params_b),
    fertility_z = standardise(normalized_fertility),
    typological_distance_z = standardise(typological_distance_from_english),
    log_resource_z = standardise(log10_common_crawl_pages),
    log_params_z = standardise(log2_params),
    # base_model is the tokenizer/weights identity shared across reasoning modes;
    # reasoning is the on/off knob (non-reasoning models are "off"). Used below.
    base_model = str_remove(model, "\\s*\\(reasoning (on|off)\\)$"),
    reasoning = factor(if_else(str_detect(model, "\\(reasoning on\\)$"), "on", "off"), levels = c("off", "on"))
  ) %>%
  arrange(family, model, language, template)

cells %>%
  select(fertility_z, typological_distance_z, log_resource_z, log_params_z) %>%
  summary()
```

Two of these are language-level, so their main effects rest on 7 points
regardless of the 566k instances behind them. Worth seeing the collinearity
before reading any coefficient:

```{r inspect-collinearity}
cells %>%
  select(fertility_z, typological_distance_z, log_resource_z, log_params_z) %>%
  cor() %>%
  round(2)
```

Fertility and typological distance are strongly related, so their individual
coefficients should not be read independently.

```{r inspect-features-plot, fig.height=4}
cells %>%
  distinct(model, language, normalized_fertility, typological_distance_from_english, log10_common_crawl_pages) %>%
  pivot_longer(c(normalized_fertility, typological_distance_from_english, log10_common_crawl_pages)) %>%
  ggplot(aes(value, language)) +
  geom_point(alpha = 0.3) +
  facet_wrap(~name, scales = "free_x") +
  labs(x = NULL, y = NULL, title = "Feature spread across languages") +
  theme_minimal()
```

Raw accuracy by language and size, as a sanity check on what the model should
recover.

Languages are ordered by resource level and coloured on a sequential ramp rather
than with categorical hues: resource is a magnitude, so the colour should carry
the ordering. `mako` runs dark to light through a single blue-green hue, is
monotonic in lightness (so the order survives greyscale and colour-vision
deficiency), and avoids the pale yellow end of `viridis` that disappears against
a white panel. Highest-resource language is darkest.

```{r language-order}
language_levels <- cells %>%
  distinct(language, log10_common_crawl_pages) %>%
  arrange(desc(log10_common_crawl_pages))

# Label with the resource value so the legend states the ordering explicitly.
language_labels <- sprintf("%s (%.1f)", language_levels$language, language_levels$log10_common_crawl_pages)
names(language_labels) <- language_levels$language

language_levels
```

```{r inspect-accuracy, fig.height=4.5}
cells %>%
  group_by(model, language, log2_params) %>%
  summarise(accuracy = sum(n_correct) / sum(n_total), .groups = "drop") %>%
  mutate(language = factor(language, levels = language_levels$language)) %>%
  ggplot(aes(log2_params, accuracy, colour = language)) +
  geom_point(alpha = 0.85, size = 1.8) +
  geom_smooth(se = FALSE, method = "loess", formula = y ~ x, linewidth = 0.9) +
  scale_colour_viridis_d(
    option = "mako", begin = 0.08, end = 0.82, direction = 1,
    labels = language_labels
  ) +
  # loess is unconstrained and overshoots 1.0 at the top end; clip the view
  # rather than the data so the smoother still sees every point.
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "log2 parameters (B)", y = "accuracy",
    colour = expression(Language ~ (log[10] ~ CC ~ pages)),
    title = "Transfer gap narrows with scale",
    subtitle = "Languages ordered by resource level, darkest = highest resource"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

# Model

Language features are model-by-language (fertility) or language-only
(typological distance, resource). Interacting them with size tests the claim that
scale governs how much of the transfer gap survives.

The baseline drops the language features and keeps size; the quantity of
interest is how far `sd(model:language)` shrinks between the two, i.e. how much
of the cross-lingual transfer variation the language features explain.

This document fits the model with lme4, keeping `template` and `family` as random
effects (the model as written in `analysis.md`) -- glmer handles them quickly. The
Bayesian counterpart in `transfer_brms.Rmd` makes those two fixed instead, purely
for sampling geometry (as random effects the template SD funnels and the
single-model family, Apertus, makes a flat ridge that saturates the NUTS
treedepth); `sd(model:language)`, the quantity of interest, is the same either way.

We save the prepared cells so the Bayesian document reads them directly rather than
re-running the 786k-row prep.

```{r save-cells}
cells <- cells %>% mutate(across(c(model, base_model, family, language, template), as.factor))
saveRDS(cells, file.path(out_dir, "model_cells.rds"))
```

`script` (shared vs. not) from `analysis.md` is deliberately absent. With only
`zho` and `rus` non-Latin out of 7 languages it is nearly collinear with
typological distance and not separately identifiable -- report it descriptively
instead.

## lme4

`glmer` wants `cbind(successes, failures)` rather than brms' `trials()` syntax.
The baseline drops the language features; comparing its `sd(model:language)` to the
full model's is how we read off the transfer variance the features explain.

```{r fit-lme4}
library(lme4)

random_effects <- "(1 | family / model) + (1 | template) + (1 | model:language)"

fit_baseline <- glmer(
  as.formula(paste("cbind(n_correct, n_total - n_correct) ~ log_params_z +", random_effects)),
  data = cells, family = binomial()
)
fit_full <- glmer(
  as.formula(paste(
    "cbind(n_correct, n_total - n_correct) ~",
    "(fertility_z + typological_distance_z + log_resource_z) * log_params_z +", random_effects
  )),
  data = cells, family = binomial()
)

summary(fit_full)
```

```{r inspect-lme4-variance}
VarCorr(fit_full)

sd_baseline <- attr(VarCorr(fit_baseline)$`model:language`, "stddev")[[1]]
sd_full <- attr(VarCorr(fit_full)$`model:language`, "stddev")[[1]]

tibble(
  model = c("baseline (size only)", "full (language features)"),
  sd_model_language = c(sd_baseline, sd_full),
  variance_explained = c(NA, 1 - sd_full^2 / sd_baseline^2)
)
```

## Analysis of results

KCE: the following is a work in progress

We see that size have the largest effect of parameters, with larger models performing (1.94, p<0.05) 
secondly see that higher resource languages leads to a higher performance (.699, p<0.05), but 
that the effect is less pronounced for larger models (-.210, p<0.05). 
Language with a higher typological distance from English, generally leads to worse performance (-.299, p<0.05)
and higher tokenizer fertility improved the likelihood of obtaining the correct answer (.193, p<0.05).

The interactions between tokenizer fertility and the model size (-.0576, p=.423), and
between typological distance and model size (.0332, p=.634) where not found to be significant,
indicating that effect of fertility and language distance remains a problem regardless of the size
(or at least we couldn't detect if it was influenced by it).

Beyond the individual coefficients we can ask how much of the cross-lingual transfer these
language features account for. The model-by-language random effect captures how much a model's
accuracy in a given language departs from its overall skill and the template difficulty -- the
transfer variation the fixed effects leave unexplained. Adding the three language features
shrinks this variance component from $\sigma^2 = 1.10$ (SD 1.05) in the size-only baseline to
$\sigma^2 = 0.39$ (SD 0.63) in the full model, a 64% reduction in variance (40% on the SD scale).
Roughly two-thirds of the transfer variation that remains after accounting for model skill and
template difficulty is thus captured by tokenizer fertility, typological distance and resource
level together.

Questions:

- [x] what about resourcity and typological distance
- [ ] what about reasoning
- Is it a difference in parsing or "reasoning"
- is there more of the variance that could be explained?
- there is high correlation between fertility and language distance (estionian might fix it). 
  - One potential solution it to decompose into fertility across model (fertility between), this should be colinear 
    to typoligical distance, while the model devisation from the langauge mean fertility (fertility within) would 
    be orthogonal
- add bootstrap on proportion of variance explained (for later)
- [x] compute the what languages we currently predict the transfer poorly on and see if there is any language features we need to add (what is the unexplained variance) 
  - potentially some systematicity, but hard to say with only 7 languages
    - non-linearity in resource: It seems to overreport on low-resource languages and underreport on high-resource (we could apply non-linearity to resourcity)
    - Chinese is likely an outlier (estimates of resourcity is likely incorrect for some of the models, e.g. Qwen)
    - 


Note when writing:
there are three actionable parameters, fertility, model size and reasoning, but depending these might have varying effects on downstream languages (hence the interactions). 
As such what we seek to answer which actionable levers can resource which low-resource axis [find a better term here]. For instance we show that while increase model size reduce the
effect diminish the effect languages prevalance in the training data, increasing the size does not counteract the effect of low tokenizer fertility or typological distance.

Note in limitations:
Our "resourcity / prevalence in the training data" is a proxy made from common crawl based on page count. While it is likely close to real world distributions, 
it likely has systematic bias (e.g. underrepresenting Chinese) and does not vary across model. Ideally this analysis would be caried out with models where the
training dataset distribution is well-known. 


## Exploratory: does more structure help?

Two additions the questions above point to, each refitting the full model with one
extra term.

A **resource x typological-distance** interaction. It is significant but a minor
lever -- variance explained moves 64.4% to 65.9%. The coefficient is positive, so a
high resource level *buffers* the typological-distance penalty: distance hurts less
when there is more data for the language.

```{r explore-resource-typo}
fit_rxt <- update(fit_full, . ~ . + typological_distance_z:log_resource_z)

anova(fit_full, fit_rxt)
summary(fit_rxt)$coefficients["typological_distance_z:log_resource_z", , drop = FALSE]
```

A **per-language residual** `(1 | language)`. Strongly significant, with
`sd(language)` ~ 0.42: roughly half of the transfer variation the three features
leave unexplained is systematic *per-language* structure, not model-by-language
idiosyncrasy -- i.e. there are language features still to add. The random-effect
estimates show which languages the features over- or under-predict, a starting
point for "which languages do we transfer to poorly".

```{r explore-language-residual}
fit_lang <- update(fit_full, . ~ . + (1 | language))

anova(fit_full, fit_lang)
VarCorr(fit_lang)
ranef(fit_lang)$language
```


## Reasoning

Reasoning is a model-level knob you can actually turn, so it gets its own model. The
reasoning-capable models are run both ways -- `Qwen3-8B (reasoning on)` vs
`(reasoning off)` -- a within-base paired design, so the effect is identified
*within* `base_model` (12 paired base models across Granite, Qwen3 and Qwen3.5;
Olmo-3-7B-Think is reasoning-on only, so it contributes no pair). We add `reasoning`
as a fixed effect and interact it with the language features and size, to ask not
just whether reasoning helps but whether it narrows the transfer gap.

```{r fit-reasoning}
fit_reasoning <- glmer(
  cbind(n_correct, n_total - n_correct) ~
    (fertility_z + typological_distance_z + log_resource_z) * log_params_z +
    reasoning * (fertility_z + typological_distance_z + log_resource_z + log_params_z) +
    (1 | family / base_model) + (1 | template) + (1 | model:language),
  data = cells, family = binomial()
)

round(summary(fit_reasoning)$coefficients, 4)
```

Reasoning raises accuracy (`reasoningon` ~ +0.30 logit, p ~ .01). More useful for
transfer, the resource advantage *shrinks* under reasoning
(`log_resource_z:reasoningon` ~ -0.19, p ~ .05): reasoning helps lower-resource
languages relatively more, so it is a lever that narrows the resource-driven gap.
Reasoning also interacts with size (`log_params_z:reasoningon` ~ +0.28, p ~ .02) --
larger models gain more on the logit scale, though small models can gain more in raw
accuracy because large models sit near the ceiling. The fertility and
typological-distance interactions with reasoning are entangled by their mutual
collinearity, as for the main effects, and are not separately identifiable here.



### Reasoning and the transfer gap

Because reasoning is a knob rather than a property of a language, its effect on transfer
is clearest from the paired models directly -- reasoning on versus off, across the twelve
base models run both ways.

```{r reasoning-transfer-gap}
paired_bases <- cells %>%
  distinct(base_model, reasoning) %>%
  count(base_model) %>%
  filter(n == 2) %>%
  pull(base_model)

by_lang <- cells %>%
  filter(base_model %in% paired_bases) %>%
  group_by(reasoning, language) %>%
  summarise(acc = sum(n_correct) / sum(n_total), .groups = "drop") %>%
  mutate(logit = qlogis(acc))

english <- by_lang %>% filter(language == "eng") %>% select(reasoning, eng_acc = acc, eng_logit = logit)

by_lang %>%
  filter(language != "eng") %>%
  left_join(english, by = "reasoning") %>%
  group_by(reasoning) %>%
  summarise(
    mean_accuracy       = mean(acc),
    gap_accuracy_points = mean(eng_acc - acc),      # absolute: English minus target, in accuracy points
    error_odds_ratio    = mean(exp(eng_logit - logit)) # relative: times more likely to err than in English
  )
```

Turning reasoning on raises accuracy across the board (~+6 points) and shrinks the
English-to-target gap from roughly 20 to 13 accuracy points. It also shrinks on the
*relative* scale: with reasoning off a model is on average about 2.9x more likely to answer
incorrectly in a non-English language than in English, and about 2.1x with it on. This
distinction matters -- a uniform lift would narrow the absolute gap on its own, because the
accuracy scale compresses near the ceiling, while leaving that error-odds ratio unchanged.
The ratio falling means reasoning does more than raise all boats. The fitted term behind it
is the `reasoning x resource` interaction (~-0.17, p ~ .05): reasoning helps lower-resource
languages somewhat more even on the log-odds scale, though with only seven languages this
differential is modest and we do not read it language by language.

(This is also why the variance-explained figure barely moves for reasoning and is the wrong
number to quote: reasoning defines the model configuration, so its transfer effect already
sits inside the `model:language` random term whether or not reasoning is a fixed effect, and
variance on the log-odds scale is invariant to a level shift. The gap above is the honest
measure.)



## Is transfer just about parsing the problem?

Does a model fail in a language because it cannot *read* the problem, or because it
cannot do the *maths* once it has read it? We proxy parsing by whether the response
reproduces the problem's numbers -- `all_prompt_numbers_present`, with the underlying
count `retrieved_prompt_number_count` out of `prompt_number_count`. This is a narrow,
mechanical slice of parsing (number transcription), not full comprehension.

Parsing is nearly necessary but far from sufficient: retrieving the numbers is almost
always needed to be correct, yet plenty of full parses still get the wrong answer.

```{r parse-decomposition}
cells %>%
  group_by(language) %>%
  summarise(
    parse_rate      = sum(n_parsed) / sum(n_total),         # all numbers retrieved
    accuracy        = sum(n_correct) / sum(n_total),
    acc_given_parse = sum(n_correct_parsed) / sum(n_parsed), # correct | parsed
    .groups = "drop"
  ) %>%
  arrange(accuracy)
```

The gap barely closes when we condition on parsing: even among instances where every
number was retrieved, English still beats Icelandic ~0.80 to ~0.51. Most of the
transfer gap sits *after* number retrieval.

Before trusting the by-language parse comparison, we check the number extractor is
not itself biased across languages. Every language uses the same templates with the
same numbers, so the extracted `prompt_number_count` should be language-invariant --
if the regex missed numbers in some language (a decimal comma, a spelled-out number,
non-Latin digits) that language would show a lower count regardless of the model.

```{r parse-extraction-check}
cells %>%
  group_by(language) %>%
  summarise(
    mean_prompt_numbers = sum(n_prompt_numbers) / sum(n_total),
    parse_rate          = sum(n_parsed) / sum(n_total),
    .groups = "drop"
  ) %>%
  arrange(mean_prompt_numbers)
```

The extracted counts are nearly identical (within ~0.1 numbers per template across
languages), so the metric's denominator is comparable and the parse-rate differences
are not an extraction artifact -- Icelandic's low parse rate comes with a perfectly
normal number count, i.e. it is the model's *responses* that reproduce the numbers
less often, not the extractor failing on Icelandic prompts. The one small asterisk is
Chinese, which extracts ~3% fewer (occasional Chinese numerals the Western-digit regex
misses). This check validates the prompt side only; a response-side formatting bias
(numbers written as words in some languages' answers) would need the completion text
to rule out, though the clean prompt side and the parse/accuracy correlation argue it
is minor.

To see which features act at which stage, we fit the same feature model to two
outcomes -- the numbers retrieved (binomial over the prompt's numbers), and
correctness *given* a full parse -- and compare their coefficients to the overall
model's.

```{r fit-parse-reason}
parse_formula <- function(lhs) as.formula(paste(
  lhs, "~ (fertility_z + typological_distance_z + log_resource_z) * log_params_z +",
  "(1 | family / model) + (1 | template) + (1 | model:language)"
))

fit_parse <- glmer(parse_formula("cbind(n_retrieved, n_prompt_numbers - n_retrieved)"),
                   data = filter(cells, n_prompt_numbers > 0), family = binomial())
fit_reason <- glmer(parse_formula("cbind(n_correct_parsed, n_parsed - n_correct_parsed)"),
                    data = filter(cells, n_parsed > 0), family = binomial())

lang_terms <- c("fertility_z", "typological_distance_z", "log_resource_z")
data.frame(
  overall            = fixef(fit_full)[lang_terms],
  parse              = fixef(fit_parse)[lang_terms],
  reason_given_parse = fixef(fit_reason)[lang_terms]
)
```

Reading these as log-odds on the standardised features: the resource effect, the
strongest transfer driver, is ~0.34 on parsing but ~0.76 on reasoning-given-parse --
essentially all of its ~0.74 overall effect is downstream of number retrieval.
Low-resource languages do not mainly fail to *read* the numbers; they fail to compute
the answer once they have them. Typological distance tilts the other way -- a little
stronger on parsing (~-0.39) than on reasoning (~-0.31), so structural distance
impedes reading the problem somewhat more. Fertility is roughly even across the two.

So the answer is **no**: transfer is not just parsing. The dominant, resource-driven
part of the gap is a reasoning effect -- the maths degrades in lower-resource languages
even when the model has correctly read the numbers -- which is the finding that bears
on treating the arithmetic as language-invariant.

Caveats. `all_prompt_numbers_present` captures only number *retrieval*; understanding
the problem's structure is not measured and stays in the "reasoning" bucket, so this
separates "not mere transcription failure" rather than "pure arithmetic". The parse
outcome sits near ceiling (~86%) while correctness is lower, so the two coefficient
columns are on different base rates and are best read for direction, not exact
magnitude. And the reason-given-parse model conditions on a full parse -- a selection
the random effects absorb but do not fully eliminate.


## Notes

Q: What was the issue with mediator / collider?
Q: Add a thing on ensuring that parsing is not conflated with reasoning 

